---
title: "Assessing Riparian Impervious Cover Change in Washington State"
date: "`r Sys.Date()`"
format:
  docx: 
    reference-doc: /Users/auerbdaa/O/code/quarto_template.docx

  html:
    embed-resources: true
    theme: yeti 
    code-fold: true
    toc: true
    toc-location: left
    grid:
      sidebar-width: 180px
      body-width: 1100px
      margin-width: 20px
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE, 
  warning = FALSE, message = FALSE, fig.width = 7, fig.height = 9)

library("tidyverse", quietly = T)
library("sf")
library("patchwork")
library("gt")
theme_set(theme_minimal()) 

dir_data_common <- "~/T/DFW-Team WDFW Watershed Synthesis - data_common"
epsg <- 2927 #WA state standard; NAD83(HARN)/ft
ha <- 0.404686

sf_wrias <- sf::read_sf(file.path(dir_data_common, "WR_WAT_WRIA.gdb/")) |>
  mutate(
    wria = factor(WRIA_NR),
    wria_nr_nm = paste0(str_pad(WRIA_NR,width = 2, pad = "0",),"_",WRIA_NM) |> factor()) |>
  select(WRIA_NR, WRIA_NM, wria, wria_nr_nm, acres = WRIA_AREA_ACR_QT) |> arrange(wria) |> 
  sf::st_transform(sf::st_crs(epsg))

sf_cnty <- sf::read_sf(file.path(dir_data_common, "WA_County_Boundaries/WA_County_Boundaries.shp")) |> 
  select(cnty = JURISDIC_2) |> 
  sf::st_transform(sf::st_crs(epsg))


#load(".RData")
```

# Authors

Dan Auerbach, RDE team, others

# Abstract

# Introduction

'Riparian' designates where streams and rivers meet and mingle with uplands. As with other ecological edges, this meeting place is host to dynamic physical and biological exchanges. These land-water interactions provide numerous services of value to people, leading to laws and policies meant to protect and rehabilitate riparian areas.

Administrative responsibility for these areas is distributed and often overlapping in Washington state. No single entity has statewide authority over riparian management or controls the location and size of jurisdictional 'riparian management zones' (RMZ). Within the context of U.S. treaties with sovereign tribes, various federal and state statutes as well as local municipal codes create a complex 'realized regulatory regime' for riparian and shoreline areas. These interacting frameworks have resulted in considerable variation in exactly what area constitutes "riparian" and thereby receives the associated protections in land use planning and permitting.

However, the Washington Department of Fish and Wildlife (WDFW) has issued formal guidance on a statewide RMZ standard that is grounded in available evidence of ecosystem functions. This standard is neither compulsory nor uniformly applicable, but it follows from the agency's mandate "to preserve, protect and perpetuate fish, wildlife and ecosystems while providing sustainable fish and wildlife recreational and commercial opportunities", and its responsibility to consult on rules and projects affecting habitat for both harvested and non-harvested species. Adoption of this minimum standard of protection increases the likelihood that comprehensive plans and critical area ordinances stipulated by Washington's Growth Management Act (GMA, RCW 36.70A) will foster healthy aquatic ecosystems and contribute to recovery of threatened and listed species, including Chinook salmon (*Oncorhynchus tshawytscha*). 

The [200-year Site Potential Tree Height standard (SPTH200)](https://wdfw.wa.gov/publications/02564) promulgated in WDFW management recommendations [(Rentz et al. 2020)](https://wdfw.wa.gov/publications/01988) follows from an extensive review and synthesis of scientific studies [(Quinn et al. 2020)](https://wdfw.wa.gov/publications/01987) that updated and extended earlier guidelines [(Knutson and Naef 1997)](https://wdfw.wa.gov/sites/default/files/publications/00029/wdfw00029.pdf). The premise that social, political, and economic values inform any answer to "how wide should a riparian buffer be" led to separation of the technical review and policy recommendations ([Wilhere and Quinn 2018](https://digitalrepository.unm.edu/nrj/vol58/iss2/13/), Chapman et al. 2020). Accordingly, the SPTH200 concept in the 2020 WDFW guidance - that the height of mature native trees characteristic of a location determines the lateral distance from water of the RMZ - seeks to accommodate the need for consistency and standardization in implementation while recognizing the inherent natural variation in riverscapes (Fausch et al. 2002).

But despite the well-understood importance of riparian areas and the considerable attention devoted to their study in Washington and the western United States more broadly, considerable uncertainty still surrounds basic measures of riparian condition and change at the landscape scale. Notwithstanding the consensus that riparian ecosystems have experienced "substantial" or "dramatic" changes, quantitative estimates remain elusive, and differences may hinder effective communication or setting appropriate policy objectives.

This analysis therefore addresses the need for landscape scale quantification, with a focus on imperviousness. Impervious cover is well-understood to be associated with ecohydrologic degradation and has statewide relevance across varied ecoregions. Specifically, we investigated the seemingly simple question "How much impervious surface has Washington state added to riparian areas in the 21st century?" Among other legal and economic shifts, this period follows the initial passage of the GMA in 1990 and the federal endangered species listings of several salmon populations during the 1990s, and it offers a perspective on the outcomes of recent practices with relevance to ongoing land use policy debates.

Past research has indicated likely patterns without providing a clear answer to "how much?". Jones et al. (2010) considered late 20th century land cover change (1973-2003) at relatively coarse spatial resolution (60m) within a 180m buffer of NHD watercourses in several thousand sample blocks across the continental United States. They described a nationwide monotonic increase in the average riparian "urban %" over the assessed duration, and their results showed the Puget lowlands region had the largest decrease in "natural land cover" during the "mid 80s - early 90s" period (relative both to its change in other time periods and to all other regions nationally in that period).

"Urban" or impervious cover within the Snohomish River watershed, north of Seattle, was also described by Powell et al. (2008). They estimated an increase of 255%, from 3285 ha in 1972 to 11,652 ha in 2006, based on a nuanced analysis combining manual orthophoto review with Landsat imagery. Though not focused on riparian areas, this study indicated that a rapid impervious expansion during the 1980s then slowed somewhat throughout the 1990s before returning to a faster rate after 2000. The authors noted the potential for changes in land use regulations to have influenced this trajectory, chiefly the 1990 passage of GMA, but stated a need for more detailed study of additional alternative factors before any direct attribution.

Broadening to more Puget Sound watersheds, Bartz et al. (2015) found similar dynamics in an assessment of annual changes derived from a LandTrendr per-pixel analysis of 30m imagery within 100m or 200m buffers of several distinct salmon habitat features. Within these assessed areas, they estimated that, by 2008, impervious cover in mainstem zones increased roughly 3% on average (172ha, 425ac) from 5820ha in 1986, and that tributary zones increased roughly 5.5% (118ha, 294ac) from 2,153ha. However, they also reported evidence of a "flatter" rate of increase in urban cover across these and the other nearshore, estuary, and floodplain habitat areas examined, with the timing of inflections varying considerably among habitat types and subbasins. These authors were unequivocal that a legacy of degradation continues to affect Puget Sound habitats, but they also concluded that more recent land cover change may indicate slower growth in additional impairment, perhaps associated with a combination of cultural, economic and regulatory factors.

Lazarus (2023) considered a statewide extent, using the National Land Cover Dataset (NLCD) and Cropland Data Layer (CDL) to examine the relative prevalence of vegetation and impervious cover classes within alternative riparian buffer configurations on lands stratified by public and private ownership. This analysis sought to address a holistic set of shifts among multiple land uses across Washington's varied ecoregions, but an increase in impervious-type classes was broadly apparent, estimated to total more than 4,500 ha (>11,000 acres) in the SPTH200 buffer between 2001 and 2019 (see table 16, as the combination of all 4 NLCD 'developed' classes).  

These studies demonstrate several of the factors that contribute quantitative uncertainty to the primary research question of "how much has riparian impervious cover changed?". Studies can differ due to alternative spatial dimensions of riparian buffers (e.g., from 10m to 200m), in addition to the positional error and varied resolution of the underlying hydrographic flowlines and waterbody areas. But accepting hydrographic error and variation, and given a particular buffer formulation, landscape scale estimates have also typically relied on categorical classifications of satellite imagery, usually at a pixel resolution corresponding to a 30m ground distance. Within an image at a certain point in time, substantial spatial variation is understood to occur within the 30x30m area of each pixel, prompting various methods to constrain the changes among classes based on algorithms that translate from continuous spectral values. 

The accuracy of estimates based on these methods is likely to increase where land cover is relatively homogeneous and change areas are large relative to pixel resolution. But much of the ongoing addition of impervious cover in Washington is concentrated within heterogeneous mosaics of multiple land covers and is thought to derive from numerous distinct parcels, with change extents frequently smaller than the 0.09ha (0.22 acres) of a single 900m2 pixel. In Washington, however, impervious cover change has also been estimated from high resolution change detection (HRCD) processing of aerial imagery at 1m resolution (Pierce 2015). The HRCD approach combines automated classification of change polygons with extensive manual review and verification against before-after image pairs to generate a detailed dataset of alterations. 

This analysis examines the intersection of these high resolution changes with a statewide polygon tessellation of riparian corridors derived from the SPTH200 standard. We also present a statewide assessment that uses readily available public land cover datasets. Based on these sources, we describe landscape scale patterns of impervious increases and consider concurrent changes in human population and housing, as well as the positioning of new riparian imperviousness relative to salmon distribution within river networks. 


# Methods 

## High Resolution Change Detection 

Land cover changes were measured following the basic methodology outlined in Pierce (2015), in which automated processing of sequential aerial imagery generated candidate polygons that were then manually reviewed for accuracy and categorical attribution. This process consists of 3 main phases: 1) object-based change detection (OBCD) to created polygons of similar pixels, 2) estimation of a probability of change for each of these units, and 3) a manual review and attribute assessment. 

High resolution orthoimagery produced by the National Agriculture Inventory Program (NAIP) has been collected at two to three year intervals beginning in 2006 (Maxwell et al. 2017). Mosaics of georectified tiles of these multispectral, aerial data formed the starting point for processing. This study is limited to the 2011 to 2013, 2013 to 2015, and 2015 to 2017 time windows due to their statewide coverage. The first two time periods were 0.914-m (3 ft) resolution and the fourth was 0.305-m (1 ft). 

Object-based segmentation of these images via Trimble eCognition software (versions 8.8 - 9.5.1) generated geodatabases of irregularly sized and shaped polygons determined by per-pixel spectral properties. Segmentation rule sets were constructed to distinguish potential changes down to the scale of about 1/50 hectare (0.05 acres, 242 pixels). These candidate polygons were assigned a probability of change estimated from random forest models fit to a training set of 1000 randomly sampled and analyst-scored units combined with an additional 500 confirmed change units from a directed search.

The short intervals between NAIP images tend to skew the estimated change probability distribution towards zero or "no change", but a relatively low change probability of 25% was used to distinguish “change-likely” and “change-unlikely” groups, reducing omission errors from "unlikely" polygons that contained real change. At least 1% of "unlikely" polygons were randomly sampled for further review.

Analysts used custom software to examine this "unlikely" sample and the set of all "likely" polygons (1-4% or roughly 1,000 - 40,000, depending on the location and time period). Visual comparison between image pairs confirmed changes, allowed assignment of a likely change agent, and addressed imperfections in the homogeneity of pixels within a change segment through assignment of categorical quartile values for the proportions of total change, lost tree canopy, new impervious surface, and new semi-pervious surface (e.g., new gravel roads). Assignment of the percent change categories introduced uncertainty due to accuracy (i.e., selection of 25% vs 50%) and imprecision (i.e., the true value falling within a 25% range). Finer divisions would tend to reduce the latter while increasing the former, but the inspection of millions of candidate change polygons has confirmed that this approach is logistically feasible way to generate a high quality dataset that captures many small changes at large regional scales.  


## Riparian Data Engine: Change within a riparian corridor

```{r build_rde_wria_library, include=F, eval=FALSE}
# #first DL csv
# walk(
#   #1:30,
#   31:62,
#   
#   ~httr2::request(paste0(
#     'https://api-ripariandata.wdfw.wa.gov/wrias/',.x,'/reaches/csv'
#   )) |> 
#     httr2::req_headers(`x-api-key` = key) |> 
#     #httr2::req_dry_run()
#     httr2::req_perform() |> 
#     httr2::resp_body_raw() |> 
#     readr::read_csv() |> 
#     readr::write_csv(
#       file = file.path(dir_data_common, paste0("rde/wrias/rde_wria_",stringr::str_pad(.x, 2, pad = "0"), ".csv"))
#     )
# )
```

```{r build_rch_tbl_wria, include=F, eval=FALSE}
#then build combined object: 1084192 x 40
#note deduplicating 52 shared across wrias: length(unique(rch_tbl$pid)) == 1084140
rch_tbl <- map_df(
  list.files(file.path(dir_data_common, "rde/wrias"), pattern = ".csv", full.names = T)
  ,
  ~readr::read_csv(.x) |>
    mutate(pid = as.character(PermanentIdentifier), PermanentIdentifier = NULL)
  ) |>
  rename(wria = WaterResourceInventoryAreaID) |>
  janitor::clean_names() |>
  rename_with(~str_replace(.,"acres","ac")) |>
  rename_with(~str_replace(.,"human_forestry_","forestry_")) |>
  rename_with(~str_replace(.,"human_nonforestry_","nonfstry_")) |>
  rename_with(~str_replace(.,"decrease","dec")) |>
  rename_with(~str_replace(.,"increase","inc")) |>
  rename_with(~str_replace(.,"impervious","impv")) |>
  rename_with(~str_replace(.,"semipervious","semi")) |>
  select(pid, everything(), -reach_id) |> 
  distinct(pid, .keep_all = T)

glimpse(rch_tbl)

#saveRDS(rch_tbl, "~/T/DFW-Team WDFW Watershed Synthesis - data_common/rde/wria_rch_tbl.rds")

```

```{r rde_wria, include=F, eval=FALSE}
rde_wria <- readRDS("~/T/DFW-Team WDFW Watershed Synthesis - data_common/rde/wria_rch_tbl.rds") |> 
  mutate(
    wria = factor(wria),
    tree_pct = all_tree_cover_ac / all_land_cover_ac,
    veg_pct = all_vegetated_ac / all_land_cover_ac
    ,
    swifd_any = !is.na(swifd_salmon_distribution),
    swifd_n = 1 + str_count(swifd_salmon_distribution,","),
    swifd_n_fct = case_when(
      is.na(swifd_n) ~ "none",
      swifd_n < 3 ~ "1-2 species",
      swifd_n >= 3 ~ "3 or more species"
    ),
    swifd_chin = case_when(
      str_detect(swifd_salmon_distribution, "Chin") ~ "salmon_with_chin",
      str_detect(swifd_salmon_distribution, "Chin", negate = T) ~ "salmon_not_chin",
      is.na(swifd_salmon_distribution) ~ "not_salmon"
    )
  ) 

```

To facilitate application of the SPTH200 standard, WDFW developed a geospatial web application that collects various datasets and "allows planners to evaluate the conditions of riparian ecosystems and make data-informed planning decisions". The foundation of this [Riparian Data Engine (RDE)](https://ripariandata.wdfw.wa.gov/) is a statewide polygon tessellation of RMZs generated from public hydrography (NHD) and soils (SSURGO) data. Various land cover, jurisdictional, water quality, and salmon distribution attributes have been calculated for each RMZ unit, enabling users to interactively examine a "reach table" consisting of rows of individual river network reaches. Filters and map-based exploration support interactive queries and comparison of summarized attribute values for these focal areas. For example, a user can rapidly filter for locations within a Water Resource Inventory Area (WRIA) or a county with a tree canopy cover proportion of interest (i.e., to target protection towards higher cover or rehabilitation towards lower cover). In addition, the RDE provides a means to quickly compare a single reach to the summarized attributes of its upstream and downstream "network neighborhoods". Such information can help to inform a range of conservation management decisions at the site scale [(WDFW 2024)](https://wdfw.medium.com/a-new-way-of-looking-at-washingtons-waterways-ad05d872c4ce).

**Additional info on spth/rmz tessellation creation and HRCD intersection**

The complete set of "reach tables" were compiled from the public RDE web application, comprising attribute values for more than 1 million individual units of the SPTH200-based polygon tesselation. The permanent identifiers that index rows in this structure can include a combination of riparian management zone (RMZ, both banks where double-banked), "extent of observed water" (EOW), and "floodplain" (primarily based on FEMA mapping) geometry.

Values of the HRCD-based increase in impervious cover were tabulated statewide and per-county boundary, recognizing that county geographic extents contain multiple jurisdictions and do not represent the jurisdictional boundaries of individual county governments.


```{r build_rde_cnty_library, eval=FALSE}
walk(
  URLencode(sort(sf_cnty$cnty))
  ,
  ~httr2::request(paste0(
    'https://api-ripariandata.wdfw.wa.gov/counties/',.x,'/reaches/csv'
  )) |>
    httr2::req_headers(`x-api-key` = key) |>
    #httr2::req_dry_run()
    httr2::req_perform() |>
    httr2::resp_body_raw() |>
    readr::read_csv() |>
    readr::write_csv(
      file = file.path(dir_data_common, paste0("rde/counties/rde_cnty_",.x, ".csv"))
    )
)

```

```{r build_rch_tbl_cnty, eval=FALSE}
#then build combined object: 1,091,144x 40
#NOT deduplicating PID shared across counties
#length(unique(rch_tbl_cnty$pid)) == 1084186 (46 more than wrias?)
#would either need to remove entirely or split+assign by proportion
#preferable to include and count twice in per-county stratification
#and/but unstratified statewide sum-over-counties needs distinct()

rch_tbl_cnty <- map_df(
  list.files(file.path(dir_data_common, "rde/counties"), pattern = ".csv", full.names = T)
  ,
  ~readr::read_csv(.x) |>
    mutate(
      pid = as.character(PermanentIdentifier), PermanentIdentifier = NULL,
      cnty = basename(.x) |> str_remove("rde_cnty_") |> str_remove(".csv") |> 
        str_replace_all("%20", " ")
      )
  ) |>
  rename(wria = WaterResourceInventoryAreaID) |>
  janitor::clean_names() |>
  rename_with(~str_replace(.,"acres","ac")) |>
  rename_with(~str_replace(.,"human_forestry_","forestry_")) |>
  rename_with(~str_replace(.,"human_nonforestry_","nonfstry_")) |>
  rename_with(~str_replace(.,"decrease","dec")) |>
  rename_with(~str_replace(.,"increase","inc")) |>
  rename_with(~str_replace(.,"impervious","impv")) |>
  rename_with(~str_replace(.,"semipervious","semi")) |>
  select(pid, cnty, wria, everything(), -reach_id)

glimpse(rch_tbl_cnty)

#saveRDS(rch_tbl_cnty, "~/T/DFW-Team WDFW Watershed Synthesis - data_common/rde/cnty_rch_tbl.rds")

```

```{r rde_cnty}
#for any statewide sum need: distinct(pid, .keep_all = T)
#no way in this table structure, without a 'zone' attribute', to differentiate RMZ-EOW-Floodplain
rde_cnty <- readRDS("~/T/DFW-Team WDFW Watershed Synthesis - data_common/rde/cnty_rch_tbl.rds") |> 
  mutate(
    wria = factor(wria),
    tree_pct = all_tree_cover_ac / all_land_cover_ac,
    veg_pct = all_vegetated_ac / all_land_cover_ac
    ,
    swifd_any = !is.na(swifd_salmon_distribution),
    swifd_n = 1 + str_count(swifd_salmon_distribution,","),
    swifd_n_fct = case_when(
      is.na(swifd_n) ~ "none",
      swifd_n < 3 ~ "1-2 species",
      swifd_n >= 3 ~ "3 or more species"
    ),
    swifd_chin = case_when(
      str_detect(swifd_salmon_distribution, "Chin") ~ "salmon_with_chin",
      str_detect(swifd_salmon_distribution, "Chin", negate = T) ~ "salmon_not_chin",
      is.na(swifd_salmon_distribution) ~ "not_salmon"
    )
  ) 

```


## OFM population and housing data

Annual county-level housing units were accessed from the Washington Office of Financial Management (OFM) Data and Research [April 1 Postcensal Estimates](https://ofm.wa.gov/washington-data-research/population-demographics/population-estimates/historical-estimates-april-1-population-and-housing-state-counties-and-cities).

```{r ofm_pop}
#non-numeric special symbol flag cols starting after 1970
#flag==1 is the total of unincorporated (2) and incorp (3), incorp the sum of munis (4)
ofm_pop <- suppressMessages(readxl::read_excel(
  file.path(dir_data_common, "census", "ofm_april1_postcensal_estimates_pop_1960-present.xlsx"),
  sheet = 2, skip = 2, na = "."
) |> 
  drop_na(Filter) |>
  janitor::clean_names() |> #glimpse() 
  select(filter, county, jurisdiction, contains("population")) |> 
  pivot_longer(contains("population"), values_to = "est_pop") |> 
  mutate(
    year = str_remove(name, "_") |> str_sub(2, 5), 
    name = NULL,
    across(c(year, est_pop), as.numeric)
  ))

```

```{r ofm_housing}
ofm_housing <- readxl::read_excel(
  file.path(dir_data_common, "census", "ofm_april1_postcensal_estimates_housing_1980_1990-present.xlsx"),
  sheet = 2, skip = 2, na = "."
) |> 
  drop_na(Filter) |>  #filter(Filter!=".") |> 
  janitor::clean_names() |> #glimpse() 
  # select(filter, county, jurisdiction, contains("total_housing")) |> 
  # pivot_longer(contains("total_housing"), values_to = "total_housing") |> 
  # mutate(
  #   year = str_sub(name, 2, 5), name = NULL,
  #   across(c(year, total_housing), as.numeric)
  #   )
  select(-line) |> 
  pivot_longer(contains("units"), values_to = "units") |> 
  mutate(
    year = str_sub(name, 2, 5), 
    across(c(year, units), as.numeric),
    name = case_when(
      str_detect(name, "total") ~ "total",
      str_detect(name, "one_unit") ~ "single",
      str_detect(name, "two_or_more") ~ "multi",
      str_detect(name, "special") ~ "other"
    )
  )

```

## StreamCat-processed NLCD

The [`StreamCatTools`](https://github.com/USEPA/StreamCatTools) (Weber 2025) R package was used to access the [StreamCat dataset](https://www.epa.gov/national-aquatic-resource-surveys/streamcat-dataset) (Hill et al. 2016), which delivers consistently geoprocessed attribute values for each reach unit of the original medium resolution NHDPlus. Derived from a 30m DEM and sometimes termed "1:100K", this hydrography captures most substantial watercourses and is the precursor to subsequent higher-resolution continent-wide elevation derived hydrographic datasets that include many smaller channel units (i.e., NHDPlusHR from 10m DEM and 3DHP from LiDAR). For each reach 'comid', StreamCat provides attribute values from many landscape datasets that are processed within the local contributing catchment ('`cat`') as well as a 100m riparian buffer ('`rp100`') on "stream lines and on-network NLCD water pixels". 

StreamCat includes values calculated from 8 cycles of the National Land Cover Dataset (NLCD; 2001, 2004, 2006, 2008, 2011, 2013, 2016, and 2019), which are based on random forest classification of satellite and other imagery sources [(Wickham et al. 2023)](https://www.tandfonline.com/doi/full/10.1080/15481603.2023.2181143). The StreamCat values represent the percentages of each of 16 land cover classes within each of the *cat* and *catrp100* areas of interest  (AOI), so that the sum across classes is 1 for each AOI (i.e., the *rp100* area is spatially nested, but `rp100` percentages are provided as distinct rather than proportional to the total *cat* area). Approximate acreages were calculated simply as:

$$ acres = pct_{nlcd-class} * areasqkm_{cat|catrp100} * 247.11$$

The StreamCat process and NLCD datasets have undergone thorough QAQC and been extensively reviewed since their initial development, resulting in their widespread use in various scientific studies. These data are a standardized, consistent means to investigate large-scale and cumulative patterns, but are unsuited to inferring site specific processes.

```{r objects_sf}
  
# ##flowlines with WRIA added deduplicated via 'largest=T'
# sf_flw <- sf::read_sf(file.path(dir_data_common, "NHDPlus17/NHDSnapshot/Hydrography/NHDFlowline.shp")) |>
#   sf::st_transform(crs = sf::st_crs(epsg)) |>
#   sf::st_zm() |>
#   select(COMID, FCODE, FTYPE, LENGTHKM) |>
#   rename_with(tolower) |>  
#   st_join(sf_cnty, left = F, largest = T) |> #inner join
#   st_join(sf_wrias |> select(wria, wria_nr_nm), left = F, largest = T) |> #inner join
#   mutate(comid = as.character(comid))
# saveRDS(sf_flw, file.path(dir_data_common, "NHDPlus17/sf_nhdp_wa_flw.rds"))
sf_flw <- readRDS(file.path(dir_data_common, "NHDPlus17/sf_nhdp_wa_flw.rds"))

# # ##catchment polys with WRIA added deduplicated via 'largest=T'
# sf_cat <- sf::read_sf(file.path(dir_data_common, "NHDPlus17/NHDPlusCatchment/Catchment.shp")) |>
#   sf::st_transform(crs = sf::st_crs(epsg)) |>
#   select(comid = FEATUREID, areasqkm = AreaSqKM) |>
#   st_join(sf_cnty, left = F, largest = T) |> #inner join
#   st_join(sf_wrias |> select(wria, wria_nr_nm), left = F, largest = T) |>
#   mutate(comid = as.character(comid))
# saveRDS(sf_cat, file.path(dir_data_common, "NHDPlus17/sf_nhdp_wa_cat.rds"))
sf_cat <- readRDS(file.path(dir_data_common, "NHDPlus17/sf_nhdp_wa_cat.rds"))

# # ~608 comids along county borders are assigned different (appropriately, due to geometry+'largest')
# #not splitting geometries (and attributes) so simpler is better
# #visual inspection indicates flowlines should have precedence for associating to StreamCat/NLCD
# mm <- inner_join(
#   sf_cat |> as_tibble() |> select(comid, cnty_cat=cnty),
#   sf_flw |> as_tibble() |> select(comid, cnty_flw=cnty),
#   by = "comid"
#   ) |> 
#   filter(cnty_cat!=cnty_flw)
# x <- sf_cat |> semi_join(mm, by = "comid")
# y <- sf_flw |> semi_join(mm, by = "comid")
# mapview::mapview(list(cat = x, flw = y, cnty = sf_cnty))

# distinct(as_tibble(sf_flw), comid) #66116
# distinct(as_tibble(sf_cat), comid) #57306
# #flowlines with no cat are most pipelines & canalditch (~3K), a third of coastline
# #but also many scattered streamriver and artificalpath statewide but large concentrations in yakima region
# #cat with no flowlines are edge cases, a pothole depressional area SE of Okanagon plus a few other scattered Eastside

```

```{r objects_streamcat_nlcd}
## recall riparian buffers are 100m, per https://www.epa.gov/national-aquatic-resource-surveys/streamcat-dataset-readme 
 
# #pull all 4 AOI, including abs areasqkm
# sc_wa_nlcd <- map(
#   set_names(c(2001, 2004, 2006, 2008, 2011, 2013, 2016, 2019))
#   ,
#   ~StreamCatTools::sc_nlcd(
#     state = "WA", year = .x,
#     aoi = 'cat,catrp100,ws,wsrp100',
#     showAreaSqKm = T
#   ) |>
#     rename_with(tolower) |> as_tibble() |>
#     mutate(comid = as.character(comid), nlcd_year = .x)
# )
# saveRDS(sc_wa_nlcd, file.path(dir_data_common, "streamcat", "sc_wa_nlcd.rds"))

# #removing 7 massive cross-border COMIDs mostly in BC; 6 are >1000km2, last is >350km2
# sf_cat |> arrange(desc(areasqkm)) |> filter(areasqkm>1000) |> ggplot() + geom_sf() + geom_sf_text(aes(label = comid))
# sf_cat |> filter(between(areasqkm,100,1000)) |> ggplot() + geom_sf(aes(fill=areasqkm)) + geom_sf_text(aes(label = comid))
# filter(sf_cat, comid =="23031682")->x; mapview::mapview(x)
comids_drop_BC <- filter(as_tibble(sf_cat), areasqkm>1000) |> distinct(comid) |> bind_rows(tibble(comid = "23031682"))

#starting object with cat and catrp100 (no Ws)
#inner_join against sf_cat to add WRIA 
#then second inner_join against sf_flw to add County and get common set across all 3
#dropping remaining comids with NA rp100
#converting pcts to 0-1
#appending 'ac[res]' cols as AOI-specific product of pct and areasqkm
sc_nlcd <- readRDS(file.path(dir_data_common, "streamcat", "sc_wa_nlcd.rds")) |> 
  map(
    ~.x |> 
      rename_with(
        ~str_remove(.,paste(c(2001, 2004, 2006, 2008, 2011, 2013, 2016, 2019),collapse = "|")) |> 
          str_remove("pct")
        ) |> 
      select(comid, nlcd_year, 
             catareasqkm, catareasqkmrp100,
             ends_with("cat"), ends_with("catrp100"))
      ) |> 
  list_rbind() |> 
  anti_join(comids_drop_BC, by = "comid") |> 
  inner_join(as_tibble(sf_cat) |> select(comid, wria, wria_nr_nm), by = "comid") |> 
  inner_join(as_tibble(sf_flw) |> distinct(comid, cnty), by = "comid") |> 
  drop_na(urbmdcatrp100) |> #single lu class captures all 231 comids NA for rp100 vals
  mutate(
    across(c(ends_with("cat"), ends_with("catrp100")), ~./100)
    ,
    # urb_cat = urbhicat + urbmdcat,
    # urb_catrp100 = urbhicatrp100 + urbmdcatrp100
    ,
    across(ends_with("cat"), list(ac = ~.* catareasqkm * 247.11)),
    across(ends_with("catrp100"), list(ac = ~.* catareasqkmrp100 * 247.11))
    ) 

sc_nlcd <- sc_nlcd |> 
  select(
    wria, wria_nr_nm, cnty,
    comid, nlcd_year, 
    catareasqkm, catareasqkmrp100,
    all_of(sort(grep("cat$", colnames(sc_nlcd), value = T))),
    all_of(sort(grep("catrp100$", colnames(sc_nlcd), value = T))),
    all_of(sort(grep("cat_ac$", colnames(sc_nlcd), value = T))),
    all_of(sort(grep("catrp100_ac$", colnames(sc_nlcd), value = T))),
)

#bring both sf to same 56879
sf_flw <- sf_flw |> inner_join(distinct(sc_nlcd, comid), by = "comid") 
sf_cat <- sf_cat |> inner_join(distinct(sc_nlcd, comid), by = "comid")

pal_nlcd <- set_names(
  c("brown","darkgreen","yellow","lightgreen","yellowgreen",
    "yellow","seagreen","white","green","blue",
    "tan","grey10","grey60","grey30","grey80","seagreen")
  ,
  str_remove(sort(grep("cat$", colnames(sc_nlcd), value = T)), "cat") 
)    

```


# Results

```{r wa_imp_statewide_summary}
# #if Lazarus ~580-720ac/y for 2001:2019
# expand_grid(ac = c(11:13)*1000, y = 18:19) |> mutate(ac_y = ac/y)

wa_imp <- rde_cnty |> 
  distinct(pid, .keep_all = T) |> 
  summarise(
    across(
      c(reach_ac, 
        rmz_ac, eow_ac, flood_ac, 
        nonfstry_impv_inc_ac
        ),
      ~sum(., na.rm = T)
    )
    , .by = swifd_chin
  ) |> 
  bind_rows(
    rde_cnty |> 
      distinct(pid, .keep_all = T) |> 
      summarise(across(c(reach_ac, rmz_ac, eow_ac, flood_ac, nonfstry_impv_inc_ac), ~sum(., na.rm = T))) |> 
      mutate(swifd_chin = "total")
  ) |> 
  rename_with(~str_remove(., "nonfstry_")) |> 
  mutate(
    reach_ha = ha*reach_ac,
    impv_inc_ha = ha*impv_inc_ac,
    impv_reach_pct = impv_inc_ac / reach_ac,
    #per KP, HRCD 2011:2017 NAIP is summer to summer
    ha_y = ha*(impv_inc_ac / 6),
    ac_y = impv_inc_ac / 6 
  )
```

```{r rde_cnty_sum}
rde_cnty_sum <- rde_cnty |> 
  summarise(
    across(
      c(reach_ac, 
        riparian_ac, rmz_ac, eow_ac, flood_ac, 
        nonfstry_impv_inc_ac, nonfstry_tree_dec_ac),
      ~sum(., na.rm = T)
    )
    , .by = cnty
  ) |> 
  mutate(
    across(
      starts_with("nonf"),
      list(
        #per_100ac = ~100*(./reach_ac)
        pct_rch = ~(./reach_ac)
      )
    )
  ) |> 
  left_join(
    rde_cnty |> 
      summarise(
        across(
          c(reach_ac, 
            #riparian_ac, rmz_ac, eow_ac, flood_ac, 
            nonfstry_impv_inc_ac 
            #, nonfstry_tree_dec_ac
          ),
          ~sum(., na.rm = T)
        ),
        .by = c(cnty, swifd_chin)
      ) |> 
      pivot_wider(names_from = swifd_chin, values_from = ends_with("_ac"))
    , 
    by = "cnty"
  ) |> 
  left_join(
    ofm_pop |> 
      filter(filter == "1", year %in% c(2011, 2017)) |> 
      select(cnty = county, year, est_pop) |> 
      #arrange(cnty, year) |> summarise(across(est_pop, list(d = ~.[2] - .[1])), .by = cnty)
      pivot_wider(names_from = year, names_glue =  "{year}_{.value}", values_from = est_pop) |> 
      mutate(d_est_pop = `2017_est_pop`- `2011_est_pop`)
    ,
    by = "cnty"
  ) |> 
  left_join(
    ofm_housing |> 
      filter(filter == "1", year %in% c(2011, 2017)) |> 
      select(cnty = county, year, name, units) |> 
      #pivot_wider(names_from = name, values_from = units) |> 
      #arrange(cnty, year) |> summarise(across(total:other, list(d = ~.[2] - .[1])), .by = cnty)
      pivot_wider(names_from = year, values_from = units) |> 
      mutate(d = `2017`- `2011`) |> 
      pivot_wider(names_from = name, values_from = where(is.numeric))
    ,
    by = "cnty"
  ) |> 
  rename_with(~str_remove(., "nonfstry_"))


sf_cnty_rde <- sf_cnty |> 
  mutate(cnty_ac = st_area(geometry) |> units::set_units("acres") |> as.numeric()) |> 
  left_join(rde_cnty_sum, by = "cnty")

```

```{r sc_nlcd_year_sum}
sc_nlcd_year_sum <- sc_nlcd |> 
  summarise(
    across(
      contains("urb") & ends_with("ac"),
      list(sum = ~sum(.))
    )
    , .by = c(nlcd_year)
  ) |> 
  pivot_longer(cols = -nlcd_year, values_to = "ac") |> 
  mutate(
    aoi = if_else(str_detect(name, "rp100"), "rp100", "cat"),
    nlcd_class = str_remove(name, "cat") |> str_remove("rp100") |> str_remove("_ac_sum"),
    name = NULL,
    ha = ha*ac
    )
```

Results are presented for the full "riparian corridor" composed of the RMZ, EOW, and floodplain polygons associated with each reach unit. As noted previously, values presented relative to the stratification by county geographic boundaries reflect various combinations of federal, tribal, state, and municipal regulatory regimes, and are not due to the actions of a single management authority.

Statewide, the assessed reaches totaled `r round(wa_imp$reach_ha[4])` hectares, with `r round(wa_imp$rmz_ac[4]/wa_imp$reach_ac[4],3)`% consisting of RMZ (supp. fig 1). Variation in the NHD, soils and floodplain data underlying the polygon tessellation contributed to differences in the overall and proportional extent of the assessed riparian corridor throughout the state, both between and within county boundaries, but total reach area was an average of approximately 22% (median 17.3%, IQR 11.9% - 28.4%) of full county area (@fig-multi_map A, supp. fig 2). The largest areas of reaches associated with the distribution of Chinook salmon were in central and northern Puget Sound, with population and housing values also highest in the vicinity of Seattle, the state's largest urban area (@fig-multi_map B-D).

```{r}
#| label: fig-multi_map

#as_tibble(sf_cnty_rde) |> mutate(pct = reach_ac/cnty_ac) |> pull(pct) |> summary()

list(
  #range(rde_cnty_sum$reach_ac)
  reach_ac = sf_cnty_rde |>  
    ggplot() + geom_sf(aes(fill = ha*reach_ac), show.legend = T) + 
    #wacolors::scale_fill_wa_c("sea", reverse = T, na.value = "pink") +
    #wacolors::scale_fill_wa_c("sea", reverse = T, limits = c(25000, 500000), oob = scales::squish, na.value = "pink") +
    #wacolors::scale_fill_wa_b("sea", reverse = T, na.value = "pink") +
    scale_fill_gradient(name = "hectares", labels = scales::label_comma(), low = "#BEE7DF", high = "#212540", na.value = "pink") +
    labs(subtitle = "Total assessed riparian corridor\nRMZ+EOW+floodplain")
  ,
  # reach_ac_cnty_ac = sf_cnty_rde |> 
  #   ggplot() + geom_sf(aes(fill = reach_ac/cnty_ac), show.legend = F) + 
  #   #scale_fill_gradient(low = alpha("purple", 0.2), high = "purple", na.value = "pink") +
  #   scale_fill_gradient(low = "#BEE7DF", high = "#212540", na.value = "pink") +
  #   labs(subtitle = "Riparian corridor area / county area")
  # ,
   
  swifd_chin = sf_cnty_rde |>  
    ggplot() + geom_sf(aes(fill = ha*reach_ac_salmon_with_chin), show.legend = T) + 
    scale_fill_gradient(name = "hectares", low = alpha("salmon", 0.2), high = "salmon") +
    labs(subtitle = "Riparian + SWIFD Chinook")
  ,
  # swifd_chin_cnty = sf_cnty_rde |>  
  #   ggplot() + geom_sf(aes(fill = reach_ac_salmon_with_chin/cnty_ac), show.legend = F) + 
  #   scale_fill_gradient(low = alpha("salmon", 0.2), high = "salmon", na.value = "pink") +
  #   labs(subtitle = "Chinook / county area")
  # ,
  
  est_pop = sf_cnty_rde |>  
    ggplot() + geom_sf(aes(fill = `2011_est_pop`), show.legend = T) + 
    scale_fill_gradient(low = alpha("orange", 0.2), high = "orange", na.value = "pink") +
    labs(fill = NULL, subtitle = "2011 estimated population")
  ,
  # est_pop_cnty = sf_cnty_rde |>  
  #   ggplot() + geom_sf(aes(fill = `2011_est_pop`/cnty_ac), show.legend = F) + 
  #   scale_fill_gradient(low = alpha("orange", 0.2), high = "orange", na.value = "pink") +
  #   labs(subtitle = "Est. pop. / county area")
  # ,
  
  housing = sf_cnty_rde |>  
    ggplot() + geom_sf(aes(fill = `2011_total`), show.legend = T) + 
    scale_fill_gradient(low = alpha("purple", 0.2), high = "purple", na.value = "pink") +
    labs(fill = NULL, subtitle = "2011 total housing units")
  # ,
  # housing_cnty = sf_cnty_rde |>  
  #   ggplot() + geom_sf(aes(fill = `2011_total`/cnty_ac), show.legend = F) + 
  #   scale_fill_gradient(low = alpha("purple", 0.2), high = "purple", na.value = "pink") +
  #   labs(subtitle = "Housing / county area")
) |> 
  wrap_plots(ncol = 2) +
  plot_annotation(tag_levels = "A")

```

During the 2011-2017 period in these HRCD data, an estimated `r round(wa_imp$impv_inc_ha[4])` hectares of new impervious cover were added to the riparian corridor (table 1; `r round(wa_imp$impv_inc_ac[4])` acres), corresponding to an annual increase of `r round(wa_imp$ha_y[4])` ha/year (`r round(wa_imp$ac_y[4])` acres/year). 

The additional area of impervious cover in non-salmon reaches was similar to that of reaches associated with Chinook distribution, but this new impervious area was an order of magnitude higher percentage of the assessed riparian corridor in Chinook-associated reaches (0.1% vs 0.01%, @tab-statewide, @fig-statewide_swifd).

```{r}
#| label: tab-statewide

wa_imp |> 
  gt(caption = "2011-17 HRCD Impervious Increase within RDE Extent") |> 
#  cols_hide(ends_with("_ac")) |> 
  # gt::grand_summary_rows(
  #   columns = c(ends_with("_ac"), ends_with("_y")), 
  #   fns = list("total") ~ sum(.),
  #   fmt = ~fmt_number(., decimals = 0)) |> 
  fmt_number(decimals = 0) |> 
  fmt_percent(columns = "impv_reach_pct")
  
```

```{r}
#| label: fig-statewide_swifd

rde_cnty_swifd <- rde_cnty |> 
  distinct(pid, .keep_all = T) |> 
  summarise(
    across(
      c(reach_ac, riparian_ac, rmz_ac, eow_ac, flood_ac, nonfstry_impv_inc_ac, nonfstry_tree_dec_ac),
      ~sum(., na.rm = T)
    ),
    #.by = swifd_any
    .by = swifd_chin
  ) |> 
  mutate(
    across(
      starts_with("nonf"),
      list(
        #per_100ac = ~100*(./reach_ac)
        pct_rch = ~(./reach_ac)
      )
    )
  )

pal_swifd_chin <- c("not_salmon" = "tan", "salmon_not_chin" = alpha("salmon", 0.5), "salmon_with_chin" = "salmon")

# {
#   rde_cnty_swifd |> 
#     ggplot() +
#     geom_col(aes(
#       swifd_chin,
#       nonfstry_impv_inc_ac,
#       fill = swifd_chin),
#       show.legend = F) + 
#     scale_fill_manual(values = pal_swifd_chin) +
#     labs(x = "SWIFD status") 
# }/{
  rde_cnty_swifd |> 
    ggplot(aes(swifd_chin, nonfstry_impv_inc_ac_pct_rch)) +
    geom_col(aes(fill = swifd_chin), show.legend = F) + 
    geom_text(
      aes(label = paste('total ha.:', round(ha*nonfstry_impv_inc_ac)))
      ,nudge_y = -0.0001
    ) +
    scale_fill_manual(values = pal_swifd_chin) +
    scale_y_continuous(labels = scales::label_percent()) +
    labs(x = "SWIFD status", y = "Impervious increase proportion of riparian corridor") 
#}

```

Most individual change areas intersecting the riparian corridor were small, with nearly 95% less than 0.4ha (1ac) and a median increase of `r round(ha*median(filter(distinct(rde_cnty, pid, .keep_all=T), nonfstry_impv_inc_ac > 0)$nonfstry_impv_inc_ac),4)` ha, considerably less than the 900m^2 area of a single 30m pixel. This pattern was broadly consistent across portions of the state with more and less total increase (@fig-impv_inc_per_rch_boxes). 

```{r statewide_size_distrib, echo=FALSE, eval=FALSE}
rde_cnty |> 
  distinct(pid, .keep_all = T) |> 
  filter(nonfstry_impv_inc_ac > 0) |> #7541
  summarise(
    across(
      nonfstry_impv_inc_ac,
      list(
        q25 = ~quantile(., p = 0.25),
        q50 = ~median(.),
        q75 = ~quantile(., p = 0.75),
        pctless1ac = ~sum(.<1)/length(.)
        )
    )
  ) |> 
  pivot_longer(everything())

```

```{r}
#| label: fig-impv_inc_per_rch_boxes

# rde_cnty |> 
#   filter(nonfstry_impv_inc_ac > 0) |> 
#   ggplot(aes(
#     nonfstry_impv_inc_ac,
#     fct_reorder(cnty, nonfstry_impv_inc_ac)
#     )) +
#   geom_jitter(height = 0.2, alpha = 0.5, size = 0.4) +
#   #geom_boxplot(varwidth = T, outliers = F, fill = NA) + 
#   #wacolors::scale_fill_wa_c("volcano", reverse = T, na.value = "pink", aesthetics = c("fill","color")) +
#   labs(y = NULL, x = "acres")

{
rde_cnty |> 
  filter(nonfstry_impv_inc_ac > 0) |> #7541
  ggplot(aes(ha*nonfstry_impv_inc_ac)) +
  #geom_vline(xintercept = 0.09, linetype = 3) +
  stat_ecdf() +
  geom_jitter(aes(y = -0.1), height = 0.05, alpha = 0.2, size = 0.1) +
  labs(y = NULL, x = "hectares", subtitle = "Cumulative distribution of per-reach impervious cover increases")
}+{
rde_cnty |> 
  filter(nonfstry_impv_inc_ac > 0) |> #7541
  # filter(nonfstry_impv_inc_ac < 20) |> #7531
  filter(nonfstry_impv_inc_ac <= 1) |>  #7148
  ggplot(aes(
    ha*nonfstry_impv_inc_ac,
    fct_reorder(cnty, nonfstry_impv_inc_ac, sum)
    )) +
  geom_vline(xintercept = 0.09, linetype = 3) + #900m2
  geom_jitter(height = 0.2, alpha = 0.2, size = 0.2) +
  geom_boxplot(varwidth = T, outliers = F, color = "blue", linewidth = 0.3, fill = NA) + 
  labs(y = NULL, x = "hectares",
       subtitle = "Per-reach impervious cover increases up to 0.4 ha (1 acre)")
} + 
  plot_layout(heights = c(1/4,3/4)) + 
  plot_annotation( tag_levels = "A")
```

Qualitatively similar statewide estimates of riparian impervious change result from the high and medium urban NLCD cover classes (>80% and 50-80% impervious respectively) within the StreamCat 100m buffer on NHDPlus watercourses. Differencing the 2011 and 2016 imagery years yields an increase of `r 30157-28642`ha (@tab-sc_statewide) for the combination of these classes, and a simple OLS linear regression on NLCD year indicates an average annual increase of approximately 338ha/year for the sum of both classes during the full 2001-2019 period (@fig-sc_nlcd_statewide_urb_year; approx. 73ha/yr for high, 265ha/yr medium). The low and open urban NLCD classes are designated as representing 50-20% and <20% impervious respectively, and the associated total area therefore includes considerable non-impervious surface. Nonetheless, despite the modest decline in the urban open NLCD class, including these did not alter the basic pattern of overall increase since 2001 and during the focal HRCD years. 

```{r}
#| label: tab-sc_statewide

# sc_nlcd_year_sum |> 
#   mutate(across(c(ha, ac), round)) |> 
#   pivot_wider(names_from = aoi, values_from = c(ha, ac)) |> 
#   gt(groupname_col = "nlcd_year") |>
#   summary_rows(columns = starts_with("ha"), fns = list(total = ~sum(.)),
#                fmt = ~fmt_number(., decimals = 0)) |> 
#   fmt_number(columns = c(starts_with("ha"), starts_with("ac")), decimals = 0)

sc_nlcd_year_sum |> 
  mutate(across(c(ha, ac), round)) |> 
  filter(aoi == "rp100") |> select(-ac) |> 
  pivot_wider(names_from = nlcd_class, values_from = c(ha)) |> 
  mutate(
    hi_md = urbhi + urbmd,
    hi_md_lo = urbhi + urbmd + urblo
    ) |>
  select(nlcd_year, High = urbhi, Medium = urbmd, Low = urblo, Open = urbop, `H+M` = hi_md, `H+M+L` = hi_md_lo) |> 
  gt(rowname_col = "nlcd_year", caption = "Est. hectares NLCD urban cover in StreamCat 100m riparian buffer") |>
  fmt_number(columns = !contains("nlcd"), decimals = 0) |> 
  tab_style(style = cell_fill(), locations = cells_body(columns = "H+M")) |> 
  tab_style(style = cell_text(weight = 'bold'), locations = cells_body(columns = "H+M", rows = nlcd_year %in% c(2011, 2016)))

# lm(
#   ha ~ nlcd_year,
#   data = sc_nlcd_year_sum |>
#     filter(aoi == "rp100", 
#            #str_detect(nlcd_class,"hi|md")
#            # str_detect(nlcd_class,"hi")
#            str_detect(nlcd_class,"md")
#            ) |>
#     summarise(across(ha, sum), .by = nlcd_year)
#   ) |>
#   broom::tidy()
#mid range 0.9*73 + 0.65*265 ~= 238/y is still > 160/y
#but low end: 0.8*73 + 0.5*265 ~191 is awfully close

```

```{r}
#| label: fig-sc_nlcd_statewide_urb_year

sc_nlcd_year_sum |> 
  filter(aoi == "rp100") |> select(-ac) |> 
  mutate(nlcd_class = factor(nlcd_class, levels = c("urbhi","urbmd","urblo","urbop"))) |> 
  ggplot(aes(nlcd_year, ha, group = nlcd_class, fill = nlcd_class)) +
  geom_col(show.legend = F) +
  # ggpmisc::stat_poly_line(se = F, show.legend = F) +
  # #ggpmisc::stat_poly_eq(ggpmisc::use_label(c("adj.R2", "p"))) +
  # ggpmisc::stat_fit_tidy(
  #       method = "lm",
  #       label.x = "left", label.y = "top", vjust = -0.7,
  #       method.args = list(formula = y ~ x),
  #       mapping = aes(
  #         label = sprintf(
  #           "est. slope = %.0f/yr", #"Slope = %.3g\np-value = %.3g",
  #           after_stat(x_estimate)#, after_stat(x_p.value)
  #           )
  #         ) #, color = "blue"
  #     ) +
  scale_fill_manual(values = pal_nlcd) +
  scale_y_continuous(labels = scales::label_comma()) +
  coord_cartesian(clip = "off") +
  facet_wrap(~nlcd_class, scales = "free", 
             labeller = as_labeller(c('urbhi' = "A: High", 'urbmd' = 'B: Med.', 'urblo' = 'C: Low', 'urbop' = 'D: Open'))) +
  theme(strip.text = element_text(hjust = 0.1))
  
```

When aggregated by county boundaries, the largest impervious increases were concentrated in more heavily urbanized portions of the state, congruent with expectations based on the geographic distributions of population and housing. Summed change areas were greatest in proximity to the large and rapidly growing metropolitan areas of Seattle, Tacoma, and Vancouver in western Washington (King, Pierce, Clark) and Spokane and the Tri-Cities in the east (@fig-impv_inc_ha_map).

```{r}
#| label: fig-impv_inc_ha_map
 
#can add Oly and Yakima, Pasco/Richland etc. with str_detect(name)
sf_wa_city <- maps::us.cities |> 
      filter(country.etc == "WA", pop > 150000) |> 
      mutate(name = str_remove(name, " WA")) |> 
      sf::st_as_sf(coords = c("long", "lat"), crs = 4326) |> 
      sf::st_transform(epsg)

sf_cnty_rde |>
  ggplot() + geom_sf(aes(fill = ha*impv_inc_ac), color = "white", show.legend = T) +
  geom_sf(data = sf_wa_city, color = "brown", show.legend = F) +
  geom_sf_text(data = sf_wa_city, aes(label = name), color = "brown", hjust = 1.1, show.legend = F) +
  scale_fill_gradient(name = "ha", low = "grey90", high = "grey10") +
  labs(
    x = NULL, y = NULL,
    title = "Riparian impervious cover increase 2011-2017", 
    subtitle = "Sum of HRCD estimates, RMZ+EOW+floodplain") +
  theme(legend.position = "bottom",
        legend.key.height = unit(0.05, "npc"),
        legend.key.width = unit(0.15, "npc")
        )

```

Normalizing impervious increase by the total riparian corridor (i.e., summed reach area within county boundary) highlighted differences in the relative density of change (@fig-impv_inc_ac_pct_reach). King county's boundaries contained the largest aggregate increase, but those within Pierce and Clark county extents both formed a larger proportion of the assessed area. Conversely, though total increases were smaller within the Kitsap, Island, and Benton boundaries, these changes made up higher proportions of the riparian corridor.

```{r}
#| label: fig-impv_inc_ac_pct_reach

{
rde_cnty_sum |> 
  ggplot(aes(ha*reach_ac, ha*impv_inc_ac)) +
  geom_point(size = 0.3, show.legend = F) + 
  geom_text(aes(label = cnty), size = 3, check_overlap = T,
            hjust = -0.1, vjust = 0,
            show.legend = F) + 
    scale_x_continuous(labels = scales::label_comma()) +
  labs(
    x = "Summed riparian corridor (ha)", 
    y = "Summed impervious increase (ha)",
    title = "Impervious increases aggregated to county boundaries") 
}/{
  rde_cnty_sum |> 
  #ggplot(aes( (ha*impv_inc_ac)/(ha*reach_ac), cnty)) +
  ggplot(aes( impv_inc_ac_pct_rch, fct_reorder(cnty, impv_inc_ac_pct_rch))) +
  geom_point(aes(size = ha*impv_inc_ac), show.legend = F) + 
    scale_size_area(max_size = 4) +
    scale_x_continuous(labels = scales::label_percent()) +
  labs(
    x = "Percentage of assessed riparian corridor",
    y = NULL) 
} +
  plot_annotation(tag_levels = "A")

```

Population increases during 2011-2017 were positively correlated with the normalized impervious increase in the riparian corridor, but considerable variation was evident among county-boundary aggregates. For example, the King and Clark areas had population gains greater than that within the Pierce boundary, but smaller proportional impervious increases. Conversely, the Kitsap extent had the fourth largest normalized impervious increase despite less population growth than many other areas (@fig-impv_inc_pop_change). A comparable pattern was evident for changes in housing units, with substantial variation around a generally positive correlation (@fig-impv_inc_housing_change). The King and Island county extents were notable for the, respectively, large and small housing gains relative to proportional riparian corridor impervious additions. 

```{r}
#| label: fig-impv_inc_pop_change

# rde_cnty_sum |>
#   ggplot(aes(impv_inc_ac_pct_rch, log10(d_est_pop))) +
#   geom_point(aes(size = `2011_est_pop`, color = `2011_est_pop`)) +
#   geom_text(aes(label = cnty), hjust = 1.1, size = 3, check_overlap = T) +
#   scale_size_area(max_size = 4, guide = "none") +
#   scale_color_gradient(name = "2011 est. pop. x1000", low = alpha("orange", 0.6), high = "brown", labels = scales::label_number(scale = 1/1000)) +
#   scale_y_continuous(name = "log10 population change 2011-2017", labels = scales::label_log(digits = 1)) +
#   scale_x_continuous(name = "Percentage of assessed riparian corridor", labels = scales::label_percent()) +
#   theme(
#     legend.position = "bottom", legend.key.width = unit(0.1, "npc")
#   )

rde_cnty_sum |> 
  #ggplot(aes(d_est_pop/`2011_est_pop`, impv_inc_ac_pct_rch)) +
  # scale_x_continuous(name = "population change 2011-2017", labels = scales::label_percent()) +
  # scale_y_continuous(name = "Percentage of assessed riparian corridor", labels = scales::label_percent())
  ggplot(aes(impv_inc_ac_pct_rch, d_est_pop/`2011_est_pop`)) +
  geom_point(aes(size = `2011_est_pop`, color = `2011_est_pop`)) +
  geom_text(aes(label = cnty), hjust = -0.1, size = 3, check_overlap = T) +
  scale_size_area(max_size = 4, guide = "none") +
  scale_color_gradient(
    name = "2011 est. pop. x1000",
    low = alpha("orange", 0.6), high = "brown", labels = scales::label_number(scale = 1/1000)) +
  scale_y_continuous(name = "Population change 2011-2017", labels = scales::label_percent()) +
  scale_x_continuous(name = "Percentage of assessed riparian corridor", labels = scales::label_percent()) +
  coord_cartesian(xlim = c(-0.0001, 0.002), ylim = c(-0.01, .13), expand = F) +
  theme(
    legend.position = "bottom", legend.key.width = unit(0.1, "npc")
  )
  
```

```{r}
#| label: fig-impv_inc_housing_change

rde_cnty_sum |>
  # select(cnty, impv_inc_ac_pct_rch, d_total, d_single, d_multi) |> 
  # pivot_longer(starts_with("d_")) |> 
  # ggplot(aes(impv_inc_ac_pct_rch, value, color = name)) +
  # facet_wrap(~name, ncol = 1, scales = "free")
  ggplot(aes(impv_inc_ac_pct_rch, d_total)) +
  geom_point(color = "purple", size = 0.8) + # aes(size = `2011_total`)
  geom_text(aes(label = cnty), hjust = 1.1, size = 3, check_overlap = T) +
  scale_x_continuous(name = "Percentage of assessed riparian corridor", labels = scales::label_percent()) +
  scale_y_continuous(name = "Change in total housing units", labels = scales::label_comma()) 

```

# Discussion

We analyzed high resolution land cover change data within an ecologically-grounded riparian corridor to quantify the magnitude of new impervious area in Washington state. A statewide estimate of `r round(wa_imp$impv_inc_ha[4])`ha (`r round(wa_imp$impv_inc_ac[4])` acres) during 2011 to 2017 represented `r round(100*wa_imp$impv_reach_pct[4], 3)`% of the total assessed extent. This estimate might be viewed as both a small portion of the entire riparian corridor and a large new addition to a sensitive portion of watersheds already subject to extensive alteration.

Riparian buffer dimensions that define the jurisdictional extent of management regimes far from uniform, with a range of legally stipulated values in state regulations and local critical area ordinances (Folkerts et al. 2020). This legal variation is in addition to the heterogeneity in realized on-the-ground conditions within any particular management domain (Swartz et al 2024). The 200-year Site Potential Tree Height standard in WDFW guidance has policy influence as a formulation of best available science, but it does not necessarily determine the footprint of protected areas. 

The full riparian corridor examined in this analysis does not represent any specific authority, and it is broader than many current management zones. These results may therefore over-estimate new impervious cover relative to the highly varied mosaic of currently binding buffer distances. Regardless of legal circumstances, however, the assessed corridor has hydrologic and ecological significance to in-channel conditions, and the RDE RMZ tessellation is based on hydrography (NHD) that is missing extensive lengths of the smaller, sometimes intermittent or ephemeral channels that contribute significantly to water quantity and quality (Brinkerhoff et al 2024). These factors increase the likelihood that this analysis under-estimates the overall functional impacts to aquatic ecosystems. 

In addition to the uncertainty in hydrography and site index values on which SPTH200 dimensions are based, some residual error persists in the high resolution change polygons despite their manual review. These sources combine to prevent a straightforward estimate of the full uncertainty in the presented values. However, convergence of the independent RDE-HRCD and StreamCat-NLCD approaches presented here, as well as the estimates developed by Lazarus (2023), lends confidence to an estimate in the hundreds of hectares a year. Even imposing an arbitrary 25% reduction still produces values on this order of magnitude. This raises questions about whether regulation and permitting subject to a project-specific "no net loss" standard can have achieved the preservation of function at a cumulative scale. If hundreds of acres of impervious cover are not being removed each year, what actions have offset those that are being added?

Moreover, the addition of upland imperviousness outside of this or any other assessed corridor from 10m to 100m has some effect on riparian and in-channel conditions (e.g., a 30m "buffer" is unlikely to have the same function within a primarily forested catchment as it does as a distinct strip adjacent to converted land). Beyond site-specific slopes and soils, the surrounding watershed context for data collection and analysis has inevitably shaped our understanding of the functional dimensions that underlie buffer formulations. The applicability of distance-decay curves likely warrants review as increasing upland imperviousness continues to alter flow paths away from saturation-excess and towards infiltration-excess dynamics. While an SPTH standard allows for some flexibility to avoid prescribing a single critical area dimension, the basic concept may be invalidated where upland surface alteration has rendered even some simple multiple of SPTH (2x or 5x etc.) insufficient to perform all desired functions.

This comparison indicates only general patterns in the configuration of land cover conversion, with various anthropogenic changes such as roads, warehouses, and other commercial spaces combined with those due to residential construction. Moreover, the summarization of changes within county boundary extents did not generate replicates suited to statistical contrast without additional longitudinal observations; as noted, these "counties" included multiple authorities, had varying hydrographic error, and are affected by distinct socio-ecological histories. Nonetheless, the variation among them in the relationships between area-normalized impervious increases, population, and housing suggests the potential for growth to come at different costs in terms of riparian condition and function (@fig-impv_inc_pop_change). Lambert et al. (2025), offered a related empirical demonstration of the alternative outcomes for habitat from policies shaping land use, showing that similar housing gains in western Washington census block groups were accompanied by substantially different amounts of tree canopy loss. Further study is needed to better disentangle the complex set of interacting factors that likely contributed to the range in riparian impervious increases accompanying similar population changes.


## Next steps

  - more on SWIFD/Chinook?

  - 3DHP RMZs from geomorphically informed SPTH200 
  
  - Stormwater, 6PPDQ (coho)

  - [task force recs](https://ofm.wa.gov/sites/default/files/public/publications/RiparianTaskforcePreliminaryRecommendationsMay2024.pdf)

  - Add county level population change numbers? fit: urb_ac_rp100 ~ year + pop + housing
  - capture that per-person increases decreasing even as abs continues to increase? or just use bookend pop numbers [now seeing this was done in Powell et al (fig 6, continuously) and Bartz et al (text, bookends)]
 

see sc_ripcorr_nlcd for initial lit notes

Hepinstall-Cymerman, J., S. Coe, and L.R. Hutyra. 2013. Urban growth patterns and growth management boundaries in the central Puget Sound, Washington, 1986–2007. Urban Ecosystems 16:109–129.

Kennedy, R.E., Z. Yang, J. Braaten, C. Copass, N. Antonova, C. Jordan, and P. Nelson. 2015. Attribution of disturbance change agent from Landsat time-series in support of habitat monitoring in the Puget Sound region, USA. Remote Sensing of Environment 166:271-285.

[UW MS thesis]
Nayak (2023) focused on the Cedar River-Lake Washington watershed, reporting a measurable increase in impervious surface area within 54 identified riparian zones, noting that 21st-century growth is more nuanced due to zoning policies, yet significant enough to influence water quality and hydrology.

Nayak, S., 2023. Impact of Riparian Landscape Patterns on Water Quality in an Urbanizing River Basin: A Case of Cedar River-Lake Washington Watershed. Master's thesis. University of Washington. Available at: https://digital.lib.washington.edu/researchworks/bitstream/handle/1773/49961/Nayak_washington_0250O_23832.pdf?sequence=1 [Accessed 4 August 2025].


Maxwell, A.E., Strager, M.P., Warner, T.A., Ramezan, C.A., Morgan, A.N. and Pauley, C.E., 2019. Large-area, high spatial resolution land cover mapping using random forests, GEOBIA, and NAIP orthophotography: Findings and recommendations. Remote Sensing, 11(12), p.1409.
https://www.mdpi.com/2072-4292/11/12/1409

Zhang, H., Gorelick, S.M. and Zimba, P.V., 2020. Extracting impervious surface from aerial imagery using semi-automatic sampling and spectral stability. Remote Sensing, 12(3), p.506.
https://www.mdpi.com/2072-4292/12/3/506

Subedi, M.R., Portillo-Quintero, C., Kahl, S.S., McIntyre, N.E., Cox, R.D. and Perry, G., 2023. Leveraging NAIP imagery for accurate large-area land use/land cover mapping: A case study in central Texas. Photogrammetric Engineering & Remote Sensing, 89(9), pp.547-560.
https://www.ingentaconnect.com/contentone/asprs/pers/2023/00000089/00000009/art00007?crawler=true&mimetype=application/pdf


# Conclusion

Washington is a leader in careful land and water use management, striving to encourage vibrant human communities without compromising the natural systems that make them possible. Yet at a statewide scale, high resolution data indicate that hundreds of acres are still being converted to impervious surface each year within sensitive riparian corridors. Attention to this ongoing change will be critical to continued progress towards salmon recovery and other valued aspects of ecological integrity.

# Acknowledgements

# Citations

Chapman, M., Satterfield, T., & Chan, K. M. A. 2020. How value conflicts infected the science
of riparian restoration for endangered salmon habitat in America’s Pacific Northwest:
Lessons for the application of conservation science to policy. Biological Conservation,
244, 108508. https://doi.org/10.1016/j.biocon.2020.108508

Fausch, K.D., Torgersen, C.E., Baxter, C.V. and Li, H.W., 2002. Landscapes to riverscapes: bridging the gap between research and conservation of stream fishes: a continuous view of the river is needed to understand how processes interacting among scales set the context for stream fishes and their habitat. BioScience, 52(6), pp.483-498.

Hill, Ryan A., Marc H. Weber, Scott G. Leibowitz, Anthony R. Olsen, and Darren J. Thornbrugh, 2016. The Stream-Catchment (StreamCat) Dataset: A Database of Watershed Metrics for the Conterminous United States. Journal of the American Water Resources Association (JAWRA) 52:120-128. DOI: 10.1111/1752-1688.12372.

Knutson, K. L., and Naef, V. L. 1997. Management recommendations for Washington’s priority
habitats: Riparian (p. 181pp.). Washington Department of Fish and Wildlife.

Lazarus, D., 2023. Current Status of Riparian Buffers in Washington State: Public and Private Land Cover Within Riparian Management Zones Based on Best Practice Guidelines (MES thesis, Evergreen State College).

Maxwell, A. E., T. A. Warner, B. C. Vanderbilt, and C. A. Ramezan. 2017. Land cover classification and feature extraction from National Agriculture Imagery Program (NAIP) Orthoimagery: a review. Photogrammetric Engineering & Remote Sensing 83:737–747.

Pierce Jr, K.B., 2015. Accuracy optimization for high resolution object-based change detection: An example mapping regional urbanization with 1-m aerial imagery. Remote Sensing, 7(10), pp.12654-12679.

WDFW. 2024. A new way of looking at Washington’s waterways. Medium. https://wdfw.medium.com/a-new-way-of-looking-at-washingtons-waterways-ad05d872c4ce 

Weber M (2025). _StreamCatTools: Tools to Work with the StreamCat API Within R and Access the Full Suite of StreamCat and LakeCat Metrics_. R package version 0.6.0, commit d8a8cce142e7cb9e3666a2660a8315a1766e5981, <https://github.com/USEPA/StreamCatTools>.

Wilhere G and Quinn T. 2018. How Wide is Wide Enough?: Science, Values, and Law in Riparian Habitat Conservation, 58 Nat. Resources J.279, https://digitalrepository.unm.edu/nrj/vol58/iss2/13/

# Supplementary info

```{r gg_rde_cnty_zones_stacked_cols}
#showing the relative/proportional composition of 'zones'
rde_cnty_sum |>
  mutate(cnty = factor(cnty) |> fct_reorder(reach_ac)) |>
  select(cnty, rmz_ac, eow_ac, flood_ac) |>
  pivot_longer(-cnty) |>
  #filter(name == "eow_ac") |> #Grant Co really does have numerous large lakes and reservoirs
  ggplot() + geom_col(aes(value, cnty, fill = name)) +
  scale_x_continuous(labels = scales::label_comma()) +
  scale_fill_manual(name = "RDE zone", values = c("lightblue","darkgreen","green")) +
  labs(x = NULL, y = NULL)

```

```{r gg_rde_cnty_sum_reach_pct_cnty}
{
sf_cnty_rde |> 
  mutate(pct = reach_ac/cnty_ac) |> 
  ggplot() + 
  geom_sf(aes(fill = pct)) +
  scale_fill_gradient(name = "reach / cnty", low = grey(0.8), high = grey(0.2))
}/{
as_tibble(sf_cnty_rde) |> 
  mutate(pct = reach_ac/cnty_ac) |> 
  ggplot() + 
  geom_col(aes(pct, fct_reorder(cnty, pct, identity), fill = pct)) +
    scale_fill_gradient(name = "reach / cnty", low = grey(0.8), high = grey(0.2)) +
  scale_x_continuous("Proportion of county area", labels = scales::label_percent()) +
  labs(y = NULL)
} +
  plot_annotation("Assessed reach area as proportion of county boundary extent")
```

```{r gg_rde_cnty_housing_stacked_cols}
rde_cnty_sum |>
  mutate(cnty = factor(cnty) |> fct_reorder(`2011_total`)) |>
  select(cnty, `2011_single`, `2011_multi`, `2011_other`) |>
  pivot_longer(-cnty) |>
  ggplot() + geom_col(aes(value, cnty, fill = name)) +
  scale_x_continuous(labels = scales::label_comma()) +
  scale_fill_manual(name = "Housing", values = c("brown","yellow","tan")) +
  labs(x = NULL, y = NULL, title = "2011 housing composition")
```

```{r gg_impv_vs_pop}
#really potentially misleading?
rde_cnty_sum |> 
  mutate(
    pct_impv = impv_inc_ac / sum(impv_inc_ac),
    pct_pop = d_est_pop / sum(d_est_pop)
  ) |> 
  ggplot(aes(pct_pop, pct_impv)) +
  geom_abline() +
  geom_point() +
  geom_text(aes(label = cnty)) +
  labs(title = "Statewide proportion impv vs pop")
```

```{r gg_impv_vs_tree_dec}

rde_cnty_sum |> #glimpse()
  #ggplot(aes(tree_dec_ac_pct_rch, impv_inc_ac_pct_rch)) +
  ggplot(aes(impv_inc_ac_pct_rch, tree_dec_ac_pct_rch)) +
  geom_point() +
  geom_text(aes(label = cnty))

# rde_cnty |> 
#   drop_na(nonfstry_impv_inc_ac) |> 
#   distinct(pid, .keep_all = T) |>
#   filter(nonfstry_tree_dec_ac + nonfstry_impv_inc_ac > 0) |> 
#   filter(nonfstry_tree_dec_ac + nonfstry_impv_inc_ac < 2) |> 
#   ggplot(aes(nonfstry_tree_dec_ac, nonfstry_impv_inc_ac)) +
#   geom_abline(linetype = 3) +
#   #geom_point(size = 0.1, alpha = 0.3)
#   #geom_hex()
#   geom_bin2d()

```

```{r gg_rde_cnty_per_reach_imp_inc_pct_reach_ac}
#could also show as pct = nonfstry_impv_inc_ac / reach_ac
rde_cnty |> 
  filter(nonfstry_impv_inc_ac > 0) |> #7541
  ggplot(aes(
    nonfstry_impv_inc_ac / reach_ac,
    fct_reorder(cnty, nonfstry_impv_inc_ac, sum)
    )) +
  geom_jitter(height = 0.2, alpha = 0.5, size = 0.4) +
  geom_boxplot(varwidth = T, outliers = F, color = "orange", fill = NA) + 
  scale_x_continuous(labels = scales::label_percent()) +
  labs(y = NULL, x = NULL,
       title = "Per-reach impervious cover increases as percentage of reach area",
       subtitle = "Reach acres = RMZ plus EOW and floodplain where present"
       )
```

```{r proportional_scatters, echo=F, eval=FALSE}
# #housing unit density relative to proportion riparian? 
# #need to think about this - maybe better as 11_17 arrows showing change in density, maybe also against 11_17 impv_inc
# as_tibble(sf_cnty_rde) |> 
#   mutate(cnty_ac = st_area(geometry) |> units::set_units("acres") |> as.numeric()) |> 
#   select(cnty, cnty_ac, reach_ac, `2011_total`, `2011_single`, `2011_multi`, `2011_other`) |>
#   pivot_longer(-c(cnty, cnty_ac, reach_ac)) |>
#   mutate(
#     reach_cnty_ac = reach_ac/cnty_ac,
#     hu_cnty_ac = value/cnty_ac, #density relative to overall boundaries
#     hu_reach_ac = value/reach_ac, #density relative to riparian corr
#     cnty = factor(cnty) |> fct_reorder(hu_cnty_ac)
#     ) |>
#   filter(name == "2011_total") |> 
#   ggplot() +
#   geom_point(aes(reach_cnty_ac, hu_cnty_ac), size = 0.4) +
#   geom_text(aes(reach_cnty_ac, hu_cnty_ac, label = cnty), size = 3) +
#   facet_wrap(~name, scales = "free")
# 

```

```{r gg_nonfstry_impv_inc_ac_per_100ac_vs_abs, echo=F, eval=F}
{
rde_cnty_sum |> 
  ggplot(aes(impv_inc_ac, impv_inc_ac_per_100ac)) +
  geom_point(size = 0.3, show.legend = F) + 
  geom_text(aes(label = cnty), size = 3, check_overlap = F,
            #nudge_x = 3, 
            hjust = 0, vjust = 1.2,
            #position = position_dodge(width = 0.01),
            show.legend = F) + 
  #geom_smooth(method = "lm") +
      coord_cartesian(xlim = c(100, 650), ylim = c(0.02, 0.18), expand = F) +
  labs(
    x = "Total acres", y = "Acres imp. inc. per 100ac",
    title = "Impervious increases by county,\narea normalized vs absolute area") 
}/{
rde_cnty_sum |> 
  ggplot(aes(impv_inc_ac, impv_inc_ac_per_100ac)) +
  geom_point(size = 0.3, show.legend = F) + 
  geom_text(aes(label = cnty), size = 3, check_overlap = T,
            nudge_x = 0.5, hjust = 0,
            show.legend = F) + 
  #geom_smooth(method = "lm") +
  coord_cartesian(xlim = c(0,75), ylim = c(0, 0.08)) +
  labs(
    x = "Total acres", y = "Acres imp. inc. per 100ac",
    subtitle = "Counties with <75ac total increase") 
}
```

::: {.content-hidden when-format="html" when-format="docx"}

## 2001 Baseline

Cover classes are distributed spatially and proportionally as expected in 2001, with more intense urban cover types concentrated in the western portion of the state, particularly in central Puget Sound, and isolated high urban hotspots corresponding to Spokane, Yakima, and Vancouver. A sum of 'forest' cover types (conifer, deciduous, and mixed) captures the large areas of private, state, tribal, and national forest in the Olympics and Cascades as well as the eastern Blue Mountain and Colville regions.

The conifer, shrub, crop, and grass classes make up the largest total areas, with the two lower intensity urban classes (`urbop` and `urblo`, at <20% impervious and <50% impervious respectively) each more extensive than the sum of the two higher intensity classes (`urbmd` and `urbhi`, at <80% and >80% impervious). The relative proportions of the total areas within each AOI are similar across most classes, but the herbaceous and wooded wetland types (`hbwet` and `wdwet`) are relatively more prevalent in the `rp100` riparian corridor, while the higher intensity urban types are slightly less so.

The total area of the `rp100` extent makes up approximately 14.8% of the total 43.3mil acres included in these data, and almost all WRIAs fall within a range of 10-20% for 'riparian corridor / total local catchment area'.

```{r sc_nlcd_2001, include=F, eval=F}
#single year, all attrib both AOI
sc_nlcd_2001 <- sc_nlcd |> 
  filter(nlcd_year==2001) 
```

```{r ggstatewide_urb_tree, include=F, eval=F}
{ 
  inner_join(
    sf_flw,
    sc_nlcd_2001 |> select(comid, ends_with("catrp100")),
    by = "comid"
  ) |> 
    ggplot() +
    geom_sf(aes(color = urbmdcatrp100+urbhicatrp100)) +
    scale_color_gradient(low = "lightblue", high = "grey20") +
    labs(subtitle = "2001 medium + high intensity urban, rp100", color = "Urb% rp100") + theme_minimal()
}+{
  inner_join(
    sf_cat,
    sc_nlcd_2001 |> select(comid, ends_with("cat")),
    by = "comid"
  ) |> 
    ggplot() +
    geom_sf(aes(fill = conifcat+decidcat+mxfstcat), color = NA) +
    scale_fill_gradient(low = "grey90", high = "darkgreen") +
    labs(subtitle = "2001 conifer+deciduous+mixed forest, cat", color = "Tree% cat") + theme_minimal()
} + plot_layout(ncol = 1)

```

```{r gt_statewide_area_cat_rp100, include=F, eval=F}
sc_nlcd_2001 |> 
  select(comid, contains("_ac"), -contains("urb_")) |> 
  summarise(across(contains("_ac"), sum)) |> 
  pivot_longer(contains("_ac")) |> 
  mutate(
    aoi = if_else(str_detect(name, "rp100"), "rp100", "cat"),
    nlcd_class = str_remove(name, "cat") |> str_remove("rp100") |> str_remove("_ac"),
    name = NULL
    ) |> 
  pivot_wider(names_from = aoi, values_from = value) |>
#recall these acreages are still nested: 
#catareasqkm includes catareasqkmrp100
  mutate(
    d_cat_rp100 = cat - rp100,
    rp100_pct_cat = rp100 / cat, 
    across(c(cat, rp100), list(pct_tot = ~./sum(.)))
    ) |> 
  arrange(nlcd_class) |> 
  gt(caption = "2001 NLCD statewide total areas by LU class and AOI") |> 
  fmt_number(c("cat","rp100","d_cat_rp100"), decimals = 0) |> 
  fmt_percent(contains("pct"), decimals = 1) |> 
  data_color(
    columns = "nlcd_class",
    palette = pal_nlcd
    ) |> 
  grand_summary_rows(columns = c("cat","rp100","d_cat_rp100"),fns = ~sum(.), fmt = ~fmt_number(., decimals = 0)) |> 
  tab_style(style = cell_borders(sides = "right"), locations = cells_body(columns = "rp100_pct_cat"))

```

```{r ggwria_rp100pctcat, include=F, eval=F, fig.height=8}
d <- sf_wrias |> 
  left_join(
    sc_nlcd_2001 |> 
      #summarise(across(contains("areasqkm"), sum)) |> 
      summarise(across(contains("areasqkm"), sum), .by = wria_nr_nm) |> 
      mutate(pct = catareasqkmrp100/catareasqkm) |> 
      arrange(pct),
    by = "wria_nr_nm"
  ) 

{
ggplot(as_tibble(d), aes(pct)) + geom_boxplot(outliers = F) + 
  geom_jitter(aes(y = 0, size = catareasqkmrp100), position = position_jitter(height = 0.3), alpha = 0.5, show.legend = F) +
  scale_x_continuous(labels = scales::label_percent(), limits = c(0, 0.2)) +
    scale_size_area() +
  labs(
    title = "Area of 'rp100' as percentage of total 'cat' area", x = NULL, y = NULL,
    subtitle = "Points scaled to rp100 area") +
    theme(axis.text.y = element_blank())
  }/{
ggplot(d) + geom_sf(aes(fill = pct)) +
  scale_fill_continuous(labels = scales::label_percent()) +
  labs(fill = "(rp100/cat)%")
} + plot_layout(heights = c(1,3)/5)
```

## Patterns since 2001

### Statewide 

Aggregating to statewide totals, higher intensity urban cover has steadily increased within both local catchments and the 100m riparian corridor at similar rates since 2001. Annualizing from a basic OLS fit of $$ acres_{urban} \sim year_{NLCD} $$, suggests that, on average, approximately 179 and 655 acres per year of high and medium intensity urban cover have been added within 100m of WA watercourses. Not all of this increase is necessarily due to losses of natural vegetation cover. However, in conjunction with the 95 acre/year increase in the "low" class, these increases substantially exceed the 264 ac/y decrease in urban "open".

```{r sc_nlcd_2019, include=F, eval=F}
#single year, all attrib both AOI
sc_nlcd_2019 <- sc_nlcd |> 
  filter(nlcd_year==2019) 
```

```{r gt_2019_statewide_area_cat_rp100, include=F, eval=F}
sc_nlcd_2019 |> 
  select(comid, contains("_ac"), -contains("urb_")) |> 
  summarise(across(contains("_ac"), sum)) |> 
  pivot_longer(contains("_ac")) |> 
  mutate(
    aoi = if_else(str_detect(name, "rp100"), "rp100", "cat"),
    nlcd_class = str_remove(name, "cat") |> str_remove("rp100") |> str_remove("_ac"),
    name = NULL
    ) |> 
  pivot_wider(names_from = aoi, values_from = value) |>
#recall these acreages are still nested: 
#catareasqkm includes catareasqkmrp100
  mutate(
    d_cat_rp100 = cat - rp100,
    rp100_pct_cat = rp100 / cat, 
    across(c(cat, rp100), list(pct_tot = ~./sum(.)))
    ) |> 
  arrange(nlcd_class) |> 
  gt(caption = "2019 NLCD statewide total areas by LU class and AOI") |> 
  fmt_number(c("cat","rp100","d_cat_rp100"), decimals = 0) |> 
  fmt_percent(contains("pct"), decimals = 1) |> 
  data_color(
    columns = "nlcd_class",
    palette = pal_nlcd
    ) |> 
  grand_summary_rows(columns = c("cat","rp100","d_cat_rp100"),fns = ~sum(.), fmt = ~fmt_number(., decimals = 0)) |> 
  tab_style(style = cell_borders(sides = "right"), locations = cells_body(columns = "rp100_pct_cat"))

```

```{r ggstatewide_ac_urb_nlcd_year, include=F, eval=F}
sc_nlcd |> 
  summarise(
    across(
      contains("urb") & ends_with("ac"),
      list(sum = ~sum(.))
    )
    , .by = c(nlcd_year)
  ) |> 
  select(-contains("urb_")) |> 
  pivot_longer(cols = -nlcd_year, values_to = "ac") |> 
  mutate(
    aoi = if_else(str_detect(name, "rp100"), "rp100", "cat"),
    nlcd_class = str_remove(name, "cat") |> str_remove("rp100") |> str_remove("_ac_sum"),
    name = NULL
    ) |>
  pivot_wider(names_from = aoi, values_from = ac) |> 
  mutate(d_cat_rp100 = cat - rp100) |> 
  pivot_longer(c(cat, rp100, d_cat_rp100), names_to = "aoi", values_to = "ac") |> 
  filter(aoi != "cat") |> 
##confirming ggpmisc values
  #filter(nlcd_class=="urbhi", aoi =="rp100") -> d; lm(ac~nlcd_year, d) -> d; broom::tidy(d); rm(d)
  split(~nlcd_class) |> 
  map(
    ~.x |> 
      ggplot(aes(nlcd_year, ac, fill = nlcd_class)) +
      geom_col(show.legend = F) +
      #geom_smooth(method = "lm", show.legend = F) +
      ggpmisc::stat_poly_line(color = "blue", se = T, show.legend = F) +
      # ggpmisc::stat_poly_eq(ggpmisc::use_label(c("adj.R2", "p")), color = "orange") +
      ggpmisc::stat_fit_tidy(
        method = "lm",
        label.x = "right", label.y = "top", vjust = -0.7,
        method.args = list(formula = y ~ x),
        mapping = aes(
          label = sprintf(
            "est. slope = %.0f ac/yr", #"Slope = %.3g\np-value = %.3g",
            after_stat(x_estimate)#, after_stat(x_p.value)
            )
          ),
        color = "blue"
      ) +
      coord_cartesian(clip = "off") +
      scale_fill_manual(values = pal_nlcd[grep(unique(.x$nlcd_class), names(pal_nlcd))]) + 
      scale_y_continuous(labels = scales::label_comma()) +
      facet_wrap(~aoi, ncol = 2, strip.position = "top", scales = "free") +
      #facet_wrap(~aoi, ncol = 3, strip.position = "top", scales = "free") +
        labs(y = paste("acres", unique(.x$nlcd_class)), subtitle = unique(.x$nlcd_class)) +
      theme(strip.text = element_text(hjust = 0.1))
  ) |> 
  wrap_plots(ncol = 1, guides = "collect")

```

### By county

Considerable variation among watersheds is evident in the overall magnitude and relative rates of urban cover change outside of and within the riparian corridor.

```{r ggwria_ac_urb_nlcd_year, include=F, eval=F, fig.height=6}
#calc approx acres per-wria, per-aoi, per-urban-class, per-year
#then estimate annual acres change rp100 per annual acre change outside rp100 (d_cat_rp100) 
d <- sc_nlcd |> 
  summarise(
    across(
      contains("urb") & ends_with("ac"),
      list(sum = ~sum(.))
    )
    , .by = c(wria, wria_nr_nm, nlcd_year)
  ) |> 
  select(-contains("urb_")) |> 
  pivot_longer(cols = -c(wria, wria_nr_nm, nlcd_year), values_to = "ac") |> 
  mutate(
    aoi = if_else(str_detect(name, "rp100"), "rp100", "cat"),
    nlcd_class = str_remove(name, "cat") |> str_remove("rp100") |> str_remove("_ac_sum"),
    name = NULL
  ) |> 
  pivot_wider(names_from = aoi, values_from = ac) |> 
  mutate(d_cat_rp100 = cat - rp100) |> 
  pivot_longer(c(cat, rp100, d_cat_rp100), names_to = "aoi", values_to = "ac") |> 
  arrange(wria, aoi, nlcd_class, nlcd_year) |> 
  nest(.by = c(wria, wria_nr_nm, aoi, nlcd_class)) |> 
  mutate(
    lm_ac_y = map(data, ~lm(ac ~ nlcd_year, data = .x)),
    lm_ac_y_est = map_dbl(lm_ac_y, ~broom::tidy(.x) |> filter(term=="nlcd_year") |> pluck("estimate")),
    ac_2001 = map_dbl(data, ~filter(.x, nlcd_year==2001) |> pluck("ac"))
  ) |> 
  select(wria, wria_nr_nm, aoi, nlcd_class, ac_2001, lm_ac_y_est) |> 
  pivot_wider(names_from = aoi, values_from = c(ac_2001, lm_ac_y_est))

# g <- d |> 
#   filter(str_detect(nlcd_class, "hi|md")) |> 
#   ggplot(aes(label = wria_nr_nm)) +
#   # geom_col(aes(lm_ac_y_est_d_cat_rp100, fct_reorder(wria_nr_nm, lm_ac_y_est_d_cat_rp100, min))) +
#   # geom_col(aes(lm_ac_y_est_rp100, fct_reorder(wria_nr_nm, lm_ac_y_est_d_cat_rp100, min)), alpha = 0.5, fill = "red") +
#   geom_point(aes(lm_ac_y_est_d_cat_rp100, lm_ac_y_est_rp100,
#                  color = nlcd_class, size = ac_2001_rp100), 
#              show.legend = F) +
#   scale_color_manual(values = pal_nlcd[grep("urb", names(pal_nlcd))]) +
#   scale_size_area() +
#   facet_wrap(~nlcd_class, ncol = 1, scales = "free") +
#   labs(
#     title = "Estimated annual change in urban cover within vs outside rp100 corridor",
#     subtitle = "Point size scaled to 2001 acres of cover in rp100",
#     x = "Est. ac/year outside rp100",
#     y = "Est. ac/year within rp100"
#   )
# plotly::ggplotly(g)

#the ratio alone is sensitive to small denominators and mostly highlights locations with less cat-wide urb
d <- sf_wrias |> 
  left_join(
    d |> 
      filter(str_detect(nlcd_class, "hi|md")) |> 
      mutate(
        ac_y_rp100_per_d_cat = lm_ac_y_est_rp100 / lm_ac_y_est_d_cat_rp100
        #across(c(lm_ac_y_est_rp100, lm_ac_y_est_d_cat_rp100), list(per = ~. / lm_ac_y_est_d_cat_rp100))
      ) |>  
      arrange(desc(ac_y_rp100_per_d_cat))
    , by = c("wria", "wria_nr_nm")
  ) 

# {
# d |> 
#   ggplot() + 
#   geom_sf(aes(fill = lm_ac_y_est_d_cat_rp100)) +
#   wacolors::scale_fill_wa_c("vantage", midpoint = 100) +
#   facet_wrap(~nlcd_class, ncol = 1) +
#   labs(
#     subtitle = "Outside 100m riparian corridor",
#     fill = "est. ac/y"
#   )
# }

{
  d |> 
  ggplot() + 
  geom_sf(aes(fill = lm_ac_y_est_d_cat_rp100)) +
  wacolors::scale_fill_wa_c("vantage", midpoint = 100) +
  facet_wrap(~nlcd_class, ncol = 1) +
  labs(
    subtitle = "Outside 100m riparian corridor",
    fill = "est. ac/y"
  ) + theme_minimal() + theme(axis.text = element_blank())
}+{
d |> 
  #filter(nlcd_class == "urbhi") |> 
  ggplot() + 
  #geom_sf(aes(fill = ac_y_rp100_per_d_cat)) +
  geom_sf(aes(fill = lm_ac_y_est_rp100)) +
  wacolors::scale_fill_wa_c("vantage", midpoint = 10) +
  facet_wrap(~nlcd_class, ncol = 1) +
  labs(
    subtitle = "Within 100m riparian corridor",
    fill = "est. ac/y"
  ) + theme_minimal() + theme(axis.text = element_blank())
} + 
  plot_layout(nrow = 1) + 
  plot_annotation(title = "Estimated annual urban cover change 2001-2019")
```

Stratification by county organizes some of this spatial variation in NLCD urban cover classes by a relevant jurisdictional unit.

```{r sc_nlcd_ac_y_cnty, include=F, eval=F}
#calc approx acres per-wria, per-aoi, per-urban-class, per-year
#then estimate annual acres change rp100 per annual acre change outside rp100 (d_cat_rp100) 
sc_nlcd_ac_y_cnty <- sc_nlcd |> 
  summarise(
    across(
      contains("urb") & ends_with("ac"),
      list(sum = ~sum(.))
    )
    , .by = c(cnty, nlcd_year)
  ) |> 
  select(-contains("urb_")) |> 
  pivot_longer(cols = -c(cnty, nlcd_year), values_to = "ac") |> 
  mutate(
    aoi = if_else(str_detect(name, "rp100"), "rp100", "cat"),
    nlcd_class = str_remove(name, "cat") |> str_remove("rp100") |> str_remove("_ac_sum"),
    name = NULL
  ) |> 
  pivot_wider(names_from = aoi, values_from = ac) |> 
  mutate(d_cat_rp100 = cat - rp100) |> 
  pivot_longer(c(cat, rp100, d_cat_rp100), names_to = "aoi", values_to = "ac") |> 
  arrange(cnty, aoi, nlcd_class, nlcd_year) 

lm_nlcd_ac_y_cnty <- sc_nlcd_ac_y_cnty |> 
  nest(.by = c(cnty, aoi, nlcd_class)) |> 
  mutate(
    lm_ac_y = map(data, ~lm(ac ~ nlcd_year, data = .x)),
    lm_ac_y_est = map_dbl(lm_ac_y, ~broom::tidy(.x) |> filter(term=="nlcd_year") |> pluck("estimate")),
    ac_2001 = map_dbl(data, ~filter(.x, nlcd_year==2001) |> pluck("ac"))
  ) |> 
  select(cnty, aoi, nlcd_class, ac_2001, lm_ac_y_est) |> 
  pivot_wider(names_from = aoi, values_from = c(ac_2001, lm_ac_y_est))

```

Differences among counties are evident in both the overall magnitude and the relative rates between the riparian corridor values and those for the encompassing local catchment.

The estimated acres of each NLCD urban class within the `rp100` AOI of example counties illustrates how the two higher impervious cover classes (darker bars) are typically present at smaller extents but exhibit proportionally larger changes among NLCD-years (fig column-series).

```{r ggcnty_ac_y_urb_cols, include=F, eval=F}
#split|>map pattern for side by side aoi is visually overwhelming with change in time of multiple nlcd_class
#focus first here on just rp100

sc_nlcd_ac_y_cnty |> 
  filter(
    aoi == "rp100", 
    #str_detect(nlcd_class, "hi|md")
    str_detect(cnty, "akim|okane|hurst|nohom")
  ) |> 
  ggplot(aes(nlcd_year, ac, group = cnty, fill = nlcd_class)) +
  geom_col(show.legend = F) +
  scale_x_continuous(breaks = c(2001, 2004, 2006, 2008, 2011, 2013, 2016, 2019)) + #unique(sc_nlcd_ac_y_cnty$nlcd_year)
  scale_y_continuous(labels = scales::label_comma()) +
  scale_fill_manual(values = pal_nlcd[grep("urb", names(pal_nlcd))]) +
  facet_grid(rows = vars(cnty), cols = vars(nlcd_class), scales = "free_y") +
  labs(
    title = "Example counties showing rp100 acreage of NLCD urban classes",
    subtitle = "Note differences in y-axis scale among counties",
    x = NULL, y = "acres"
    ) +
  theme(axis.text = element_text(size = 7))

```

Clicking individual county names in the legend will focus on only those traces.

```{r plotly_ac_y_urb_lines_multi, include=FALSE, eval=FALSE}
lapply(
  sc_nlcd_ac_y_cnty |>
    filter(aoi == "rp100") |>
    split(~nlcd_class)
  ,
  \(x){
    plotly::plot_ly(
      data = x,
      type = 'scatter', mode = 'markers+lines',
      name = ~cnty, x = ~nlcd_year, y = ~ac,
      color = ~nlcd_class, colors = pal_nlcd[grep("urb", names(pal_nlcd))]
      # #generates a single 4 panel with per-panel names and without duplicated labels
      # #but label on/off not shared across panels 
      #showlegend = x$nlcd_class[1] == "urbhi"
    ) |>
      plotly::layout(
        annotations = list(
          x = 2001, y = 1.05*max(x$ac),
          text = x$nlcd_class[1],
          showarrow = FALSE
        )
      )
  }
) #|> plotly::subplot(nrows = 4)
```

Viewed as the linear regression estimate of per-year change, the medium and high classes show consistent increases (i.e., positive slope estimates) across counties, while the low and open classes show variation with some counties decreasing in these classes.
(fig boxplot of rp100 lm-est by nlcd-class)

```{r ggcnty_ac_y_urb_box, include=F, eval=F}
## if not html
# lm_nlcd_ac_y_cnty |> 
#   ggplot(aes(lm_ac_y_est_rp100, nlcd_class, color = nlcd_class)) +
#   geom_vline(xintercept = 0, linewidth = 0.3) +
#   geom_jitter(aes(size = ac_2001_rp100), position = position_jitter(height = 0.2), alpha = 0.5, show.legend = F) +
#   geom_boxplot(outliers = F, show.legend = F) +
#   scale_size_area(max_size = 4) +
#   scale_color_manual(values = pal_nlcd[grep("urb", names(pal_nlcd))]) +
#   labs(
#     title = "Annual changes by county in NLCD urban classes within riparian corridor",
#     subtitle = "Sample distributions of linear estimate of rp100 acres ~ NLCD year",
#     x = "est. acres/year", y = NULL
#   )

lm_nlcd_ac_y_cnty |> 
  mutate(nlcd_class = factor(nlcd_class)) |> 
  plotly::plot_ly(
        color = ~nlcd_class, colors = pal_nlcd[grep("urb", names(pal_nlcd))]
  ) |> 
  plotly::add_trace(
    x = ~lm_ac_y_est_rp100, 
    y = ~nlcd_class, 
    type = "box", hoverinfo = 'name+x', showlegend = F
  ) |> 
  plotly::add_markers(
    x = ~lm_ac_y_est_rp100,
    y = ~nlcd_class, 
    marker = list(size = 6),
    hoverinfo = 'text', text = ~paste(cnty, round(lm_ac_y_est_rp100,3)),
    showlegend = F
  )

```

These shifts within the riparian corridor are informed by the trajectory of urban cover in the area of each local catchment outside of the `rp100` extent. Comparing the medium and high classes this way, the within-corridor increases show clear inflections at roughly 30ac/y and 100ac/y for medium and high respectively, as well as saturation with increasing outside-corridor increases. Notably, the considerable scatter in within-corridor values at a given outside-corridor indicates clear differences between counties in where urban cover increases may have been concentrated.

(fig scatter of rp100 lm-est ~ d_cat_rp100 lm_est)

```{r ggcnty_ac_y_urb_scatter, include=F, eval=F}
g <- lm_nlcd_ac_y_cnty |> 
  filter(str_detect(nlcd_class, "hi|md")) |> 
  ggplot(aes(label = cnty, color = nlcd_class)) +
  geom_point(aes(lm_ac_y_est_d_cat_rp100, lm_ac_y_est_rp100,
                 size = ac_2001_rp100), 
             show.legend = F) +
  geom_smooth(aes(lm_ac_y_est_d_cat_rp100, lm_ac_y_est_rp100)) +
  scale_color_manual(values = pal_nlcd[grep("urb", names(pal_nlcd))]) +
  scale_size_area() +
  facet_wrap(~nlcd_class, ncol = 1, scales = "free") +
  labs(
    title = "Estimated annual change in urban cover within vs outside rp100 corridor",
    subtitle = "Point size scaled to 2001 acres of cover in rp100",
    x = "Est. ac/year outside rp100",
    y = "Est. ac/year within rp100"
  )

plotly::ggplotly(g)
```

As an example, focusing on the `urbhi` class, Thurston (~77ac/y) and Yakima (~72ac/y) counties exhibit similar increases in the area outside `rp100`, but the ~14ac/yr within the riparian corridor of the latter was roughly 7 times the ~2ac/y of the former. Within Yakima county, most of the riparian corridor 'urban high' increase appears to have occurred in the same local catchments that experienced appreciable non-`rp100` increase. In contrast, the Thurston county local catchment with the most urban increase outside of `rp100` had relatively little change within the corridor. However, at the aggregated county scale, the larger total `rp100` increase in Yakima county appears due to the more widespread occurrence of reaches with both larger and smaller increases. It is possible that some of these smaller increases may be related to data artifacts, perhaps related to the more arid conditions.

But...compare Snohom and Spokane, which are reversed E/W-dry/wet and below/above avg for rp100~d_cat_rp100

```{r ggcnty_ac_y_urb_sf_focal_cnty, include=F, eval=F}
d <- sc_nlcd |> 
  filter(
    str_detect(cnty, "hurst|akima")
    ,
    nlcd_year %in% c(2001, 2019)
    ) |> 
  #select(cnty, comid, nlcd_year, urbhicatrp100_ac) |> 
  #pivot_wider(names_from = nlcd_year, values_from = urbhicatrp100_ac) |> 
  #mutate(d = `2019`-`2001`) |> 

  select(cnty, comid, nlcd_year, contains("urbhi") & contains("_ac")) |> 
  mutate(
    #trace cover class percentages can generate negative d_cat_rp100 at single comid scale (e.g., 23105084)
    #so coerce rp100 down to cat value if greater
    urbhicatrp100_ac = if_else(urbhicatrp100_ac > urbhicat_ac, urbhicat_ac, urbhicatrp100_ac),
    urbhi_d_cat_rp100 = urbhicat_ac - urbhicatrp100_ac
    , urbhicat_ac = NULL
    ) |>  
  #filter(urbhi_d_cat_rp100 < 0) |> arrange(desc(abs(urbhi_d_cat_rp100)))
  pivot_wider(names_from = nlcd_year, values_from = contains("urbhi")) |> 
  mutate(
    d_rp100 = urbhicatrp100_ac_2019 - urbhicatrp100_ac_2001,
    d_out = urbhi_d_cat_rp100_2019 - urbhi_d_cat_rp100_2001
    ) |> 
  filter(urbhi_d_cat_rp100_2019 > 0 | urbhicatrp100_ac_2019 > 0) |> 
  arrange(d_out) |> 
  #12 cases of very small negative diffs in d_out for 19-01, all with d_rp100 < 1
  #could coerce to 0 but simpler to exclude
  filter(d_out >= 0)

d_sf_cat <- sf_cat |> 
  select(-cnty) |> 
  inner_join(d, by = "comid")

d_sf_flw <- sf_flw |> 
  select(-cnty) |> 
  inner_join(d, by = "comid")

{
ggplot() +
  geom_sf(data = d_sf_cat, aes(fill = d_out), color = "lightgrey", linewidth = 0.1) +
  geom_sf(data = d_sf_flw, color = "lightblue", linewidth = 0.3) +
  wacolors::scale_fill_wa_c("forest_fire", limits = c(0, max(d$d_out)*1.05), aesthetics = c("fill","color"), reverse = T) +
  ggspatial::annotation_scale() +
    labs(subtitle = "Outside rp100 corridor")
}/{
ggplot() +
  geom_sf(data = d_sf_cat, color = "lightgrey", fill = "white", linewidth = 0.1) +
  geom_sf(data = d_sf_flw, aes(fill = d_rp100, color = d_rp100)) +
  wacolors::scale_fill_wa_c("forest_fire", limits = c(0, max(d$d_rp100)*1.05), aesthetics = c("fill","color"), reverse = T) +
  ggspatial::annotation_scale() +
    labs(subtitle = "Within rp100 corridor")
}+
  plot_layout(ncol = 1) |> 
  plot_annotation(
    title = "Estimated increase in acres NLCD urban high cover class", subtitle = "Thurston and Yakima counties"
  )
```

```{r arrowswarm, include=F, eval=F}
comids_2019_urb <- sc_nlcd |> 
  filter(nlcd_year == 2019, (urbmdcat + urbhicat > 0)) |> 
  distinct(comid)

d <- sc_nlcd |> 
  filter(nlcd_year %in% c(2001, 2019)
         #, comid %in% comid_2019_urb
  ) |> 
  select(comid, nlcd_year, 
         wria, wria_nr_nm,
         cnty,
         contains("urbhi") & contains("_ac"),
         contains("urbmd") & contains("_ac")
  ) |> 
  mutate(
    #trace cover class percentages can generate negative d_cat_rp100 at single comid scale (e.g., 23105084)
    #so coerce rp100 down to cat value if greater
    urbhicatrp100_ac = if_else(urbhicatrp100_ac > urbhicat_ac, urbhicat_ac, urbhicatrp100_ac),
    urbhi_d_cat_rp100 = urbhicat_ac - urbhicatrp100_ac
    , urbhicat_ac = NULL,
    urbmdcatrp100_ac = if_else(urbmdcatrp100_ac > urbmdcat_ac, urbmdcat_ac, urbmdcatrp100_ac),
    urbmd_d_cat_rp100 = urbmdcat_ac - urbmdcatrp100_ac
    , urbmdcat_ac = NULL
    ) |>  
  pivot_longer(contains("cat"), names_to = "nlcd_class_aoi") |> 
  pivot_wider(names_from = nlcd_year, values_from = value) |> 
  #mutate(d = `2019`-`2001`) |> 
  pivot_longer(c(`2001`,`2019`)) |> 
  mutate(
    nlcd_class_aoi = paste0(nlcd_class_aoi,"_",name), name = NULL
  ) |> 
  pivot_wider(names_from = nlcd_class_aoi, values_from = value)

d |> 
    filter(
      #wria %in% wrias
      str_detect(cnty, "akim|okane|hurst|nohom")
      ) |> 
    ggplot() + 
    geom_segment(aes(
      x = urbhi_d_cat_rp100_2001, y = urbhicatrp100_ac_2001,
      xend = urbhi_d_cat_rp100_2019, yend = urbhicatrp100_ac_2019
      #,color = d_cat - d_catrp100
    ),
    arrow = arrow(length = unit(0.05,"npc")),
    show.legend = T) +
    geom_abline(slope = 1, intercept = 0, linetype = 3) +
    wacolors::scale_color_wa_c("lopez", midpoint = 0) +
    # scale_x_continuous(labels = scales::label_percent()) +
    # scale_y_continuous(labels = scales::label_percent()) +
    facet_wrap(~cnty, ncol = 1, strip.position = "right") +
    theme(strip.text = element_text(size = 5), legend.position = "top") +
    labs(
      title = "2001 to 2019 urban cover change",
      subtitle = "StreamCat NLCD urban high",
      x = "local catchment less riparian",
      y = "100m riparian corridor"
    )

```

:::
